---
title: "Posvirus Host Prediction"
author: "Katie Tseng, Dan Becker, Heather Koehler, Pilar Fernandez, Stephanie Seifert"
output:
  pdf_document:
  latex_engine: xelatex
  toc: yes
html_document:
  fig_height: 6
  fig_width: 6
  highlight: tango
  theme: journal
---

```{r, echo=F, message=F}
knitr::opts_chunk$set(eval=F)
```

# Introduction

The following code reproduces the analysis from:

> etc.

# Load required packages

```{r echo=F, message=T}

library(ape)
library(dplyr)
library(nlme)
library(tidyverse)
library(vroom) 

## treespace dependencies include XQuartz v2.7.11 and package 'rgl' (https://stackoverflow.com/a/66127391/2554330)
    #1. install XQuartz v2.7.11 (https://www.xquartz.org/releases/XQuartz-2.7.11.html); 
    #2. install package 'rgl': >install.packages ("rgl") \ >options(rgl.useNULL=TRUE) \ >.rs.restartR() \ >library(rgl)
library(treespace) 

## for modeling
library(gbm);
library(ROCR);
library(vegan);
library(plyr);library(dplyr);
library(mvtnorm)
library(xtable)
library(parallel)

```

# Set system

```{r echo=F, message=T}

#clean environment
rm(list=ls()) 
graphics.off()

#set working directory
setwd("~/Library/CloudStorage/OneDrive-WashingtonStateUniversity(email.wsu.edu)/Fernandez Lab/Projects (Active)/OPV Host Prediction/GitHub/PoxHost/Tseng2022")

```

# Load data

```{r echo=F, message=T}

##(1) load data
load("Tseng2022.RData")

##(2) poxdata: host-OPV interactions detected via PCR/isolation from Virion database; <https://github.com/viralemergence/virion/tree/main/Virion>
poxdata <- virion %>% filter(VirusGenus == "orthopoxvirus" & (DetectionMethod %in% c("PCR/Sequencing","Isolation/Observation"))) 

##(3) taxa: mammal species taxonomy from vertlife, <https://data.vertlife.org/>
taxa <- vertlife

##(4) hostTraits: mammal traits from the COMBINE database, <https://doi.org/10.1002/ecy.3344> (file: "trait_data_imputed.csv")
hostTraits <- combine

##(5) hostTree: mammal phylogeny tree from Dryad, <https://doi.org/10.5061/dryad.tb03d03> (path: "Data_S8_finalFigureFiles">"_DATA")
hostTree <- dryad

##(6) virusTraits: OPV accessory genes from ...
virusTraits <- opvgenes

##(7) clean environment
rm(virion, vertlife, dryad, combine, opvgenes)

```

# Aggregate poxdata to genus-level

```{r echo=F, message=T}

##(1) exclude if host genus or virus is NA and if variola virus
poxdata <- poxdata[!is.na(poxdata$HostGenus),]
poxdata <- poxdata[!is.na(poxdata$Virus),]
poxdata <- poxdata[!(poxdata$Virus=="variola virus"),]

##(2) to dis-aggregate West African from Congo Basin MPXV clades, subset MPXV interactions, export, and merge clade-specific data into poxdata
mpxvdata <- poxdata %>% filter(Virus=="monkeypox virus" & (DetectionMethod %in% c("PCR/Sequencing","Isolation/Observation")))
#write.csv(mpxvdata, "~/mpxvdata.csv")

##(3) merge clade-specific data

##(4) extract PCR-positive data 
pcr <- subset(poxdata[which(poxdata$DetectionMethod=="PCR/Sequencing"),], select=c("Host","HostGenus","Virus"))
pcr$Host <- ifelse(is.na(pcr$Host),"sp.",pcr$Host)
pcr$pcr <- 1
pcr <- aggregate(.~Host+HostGenus+Virus, data=pcr, sum)

##(5) extract isolation-positive data 
competence <- subset(poxdata[which(poxdata$DetectionMethod=="Isolation/Observation"),], select=c("Host","HostGenus","Virus"))
competence$Host <- ifelse(is.na(competence$Host),"sp.",competence$Host)
competence$competence <- 1
competence <- aggregate(.~Host+HostGenus+Virus, data=competence, sum)

##(6) merge PCR/isolation-positive data; create binary vars
poxdata <- merge(pcr, competence, by=c("Host","HostGenus","Virus"), all=TRUE)

##(7) combine studies into single variable
poxdata$studies <- ifelse(is.na(poxdata$pcr),0,poxdata$pcr) + ifelse(is.na(poxdata$competence),0,poxdata$competence)

##(8) create binary variables
poxdata$pcr=ifelse(is.na(poxdata$pcr),0,1)
poxdata$competence=ifelse(is.na(poxdata$competence),0,1)

##(9) aggregate at genus level
agg_pcr <- aggregate(pcr~HostGenus+Virus, data=poxdata, max)
agg_competence <- aggregate(competence~HostGenus+Virus, data=poxdata, max)
agg_studies <- aggregate(studies~HostGenus+Virus, data=poxdata, sum)

##(10) merge pcr, competence and studies variables
gdata <- merge(agg_pcr,agg_competence)
gdata <- merge(gdata,agg_studies)

##(11) rename variables
gdata <- rename(gdata,c('HostGenus'='gen','Virus'='virus'))
gdata$gen <- str_to_title(gdata$gen)

##(12) clean environment
rm(poxdata, mpxvdata, pcr,competence,agg_pcr, agg_competence, agg_studies)

```

# Merge poxdata with broader mammal taxa to create pseudoabsences

```{r echo=F, message=T}

##(1) drop duplicate genera in taxa
gtaxa <- taxa[!duplicated(taxa$gen),]
gtaxa <- gtaxa[c('gen','fam','ord')]

##(2) check for mismatched names, then merge gdata with taxa
gdata$gen[!gdata$gen %in% taxa$gen]
data=merge(gtaxa,gdata,by='gen',all.x=TRUE)


##(3) keep only genera from orders in which positive associations exist
keep=subset(data, pcr==1 | competence==1)
data$keep=ifelse(data$ord %in% keep$ord,TRUE,FALSE)
data=subset(data,keep==TRUE)
data$keep=NULL

##(4) create dataframe of all possible host-OPV combinations (for mammal genera that exist in orders w/ known OPV predictions)
uniq_gen <- unique(data$gen[!is.na(data$gen)])
uniq_virus <- unique(data$virus[!is.na(data$virus)])
combinations <- expand.grid(uniq_gen,uniq_virus)
combinations <- rename(combinations,c('Var1'='gen','Var2'='virus'))

##(5) merge host-OPV interaction data with all possible combinations
data <- merge(combinations,data,all.x=TRUE)

##(6) create binary variable for sampled
data$sampled=ifelse(is.na(data$pcr) & is.na(data$competence),0,1)

##(7) reclassify NAs as pseudo-absences for viral detection
data$pcr=ifelse(is.na(data$pcr),0,data$pcr)
data$competence=ifelse(is.na(data$competence),0,data$competence)
data$studies=ifelse(is.na(data$studies),0,data$studies)

##(8) replace NA taxonomic values based on host genera
data=merge(data,gtaxa,by='gen',all.x=TRUE)
data <- rename(data,c('fam.y'='fam','ord.y'='ord'))
data$fam.x=NULL
data$ord.x=NULL

##(9) clean environment
rm(gdata,taxa,gtaxa,keep,uniq_gen,uniq_virus,combinations)

```

# Aggregate hostTraits to genus-level

```{r echo=F, message=T}

##(1) observe variable names
colnames(hostTraits)

##(2) to aggregate continuous/integer variables, use the median as the summary measure
hostTraits_continuous=aggregate(cbind(adult_mass_g,brain_mass_g,adult_body_length_mm,adult_forearm_length_mm,
                                   max_longevity_d,maturity_d,female_maturity_d,male_maturity_d,
                                   age_first_reproduction_d,gestation_length_d,teat_number_n,
                                   litter_size_n,litters_per_year_n,interbirth_interval_d,
                                   neonate_mass_g,weaning_age_d,weaning_mass_g,generation_length_d,
                                   dispersal_km,density_n_km2,home_range_km2,social_group_n,
                                   dphy_invertebrate,dphy_vertebrate,dphy_plant,
                                   det_inv,det_vend,det_vect,det_vfish,det_vunk,det_scav,det_fruit,det_nect,det_seed,det_plantother,det_diet_breadth_n,
                                   upper_elevation_m,lower_elevation_m,altitude_breadth_m,habitat_breadth_n) 
                             ~ order+family+genus, data=hostTraits, FUN=median, na.action=na.pass, na.rm=TRUE)
# 'na.action=na.pass, na.rm=TRUE' is specified such that if species w/in a genus has a combination of real values & NAs, the median of real values will be returned (as opposed to omitting the genus or returning NA)

##(3) to aggregate binary variables, use the mean as the summary measure
hostTraits$fossoriality[hostTraits$fossoriality==2]<-0  #recode 0/1
hostTraits_binary=aggregate(cbind(hibernation_torpor,fossoriality,freshwater,marine,terrestrial_non.volant,terrestrial_volant,
                               island_dwelling,disected_by_mountains,glaciation) ~ order+family+genus, data=hostTraits, FUN=mean, na.action=na.pass, na.rm=TRUE)

##(4) to aggregate categorical variables transform into binary
hostTraits_cat <- hostTraits
hostTraits_cat$trophic_herbivores <- ifelse(hostTraits_cat$trophic_level==1,1,0)
hostTraits_cat$trophic_omnivores <- ifelse(hostTraits_cat$trophic_level==2,1,0)
hostTraits_cat$trophic_carnivores <- ifelse(hostTraits_cat$trophic_level==3,1,0)
hostTraits_cat$activity_nocturnal <- ifelse(hostTraits_cat$activity_cycle==1,1,0)
hostTraits_cat$activity_crepuscular <- ifelse(hostTraits_cat$activity_cycle==2,1,0) #nocturnal/crepuscular, cathemeral, crepuscular or diurnal/crepuscular
hostTraits_cat$activity_diurnal <- ifelse(hostTraits_cat$activity_cycle==3,1,0)
hostTraits_cat$forager_marine <- ifelse(hostTraits_cat$foraging_stratum=="M",1,0)
hostTraits_cat$forager_ground <- ifelse(hostTraits_cat$foraging_stratum=="G",1,0) 
hostTraits_cat$forager_scansorial <- ifelse(hostTraits_cat$foraging_stratum=="S",1,0)
hostTraits_cat$forager_arboreal <- ifelse(hostTraits_cat$foraging_stratum=="Ar",1,0)
hostTraits_cat$forager_aerial <- ifelse(hostTraits_cat$foraging_stratum=="A",1,0)
hostTraits_cat$island_end_marine <- ifelse(hostTraits_cat$island_endemicity=="Exclusively marine",1,0)
hostTraits_cat$island_end_mainland <- ifelse(hostTraits_cat$island_endemicity=="Occurs on mainland",1,0)
hostTraits_cat$island_end_lgbridge <- ifelse(hostTraits_cat$island_endemicity=="Occurs on large land bridge islands",1,0)
#hostTraits_cat$island_end_smbridge <- ifelse(hostTraits_cat$island_endemicity=="Occurs on small land bridge islands",1,0)
hostTraits_cat$island_end_isolated <- ifelse(hostTraits_cat$island_endemicity=="Occurs only on isolated islands",1,0)
hostTraits_cat$biogeo_afrotropical <- ifelse(grepl("Afrotropical",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_antarctic <- ifelse(grepl("Antarctic",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_australasian <- ifelse(grepl("Australasian",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_indomalayan <- ifelse(grepl("Indomalayan",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_nearctic <- ifelse(grepl("Nearctic",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_neotropical <- ifelse(grepl("Neotropical",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_oceanian <- ifelse(grepl("Oceanian",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_palearctic <- ifelse(grepl("Palearctic",hostTraits_cat$biogeographical_realm),1,0)

##(5) to aggregate transformed categorical-to-binary variables, use the mean as the summary measure
hostTraits_cat=aggregate(cbind(trophic_herbivores,trophic_omnivores,trophic_carnivores,
                            activity_nocturnal,activity_crepuscular,activity_diurnal,
                            forager_marine,forager_ground,forager_scansorial,forager_arboreal,forager_aerial,
                            island_end_marine,island_end_mainland,island_end_lgbridge,island_end_isolated,
                            biogeo_afrotropical,biogeo_antarctic,biogeo_australasian,biogeo_indomalayan,biogeo_nearctic,biogeo_neotropical,biogeo_oceanian,biogeo_palearctic)
                       ~ order+family+genus, data=hostTraits_cat, FUN=mean, na.action=na.pass, na.rm=TRUE)

##(6) merge continuous variables with binary variables and clean environment
hostTraits <- full_join(hostTraits_continuous, hostTraits_binary, by = c("order","family","genus"),keep=TRUE)
hostTraits <- rename(hostTraits,c('order.x'='order','family.x'='family','genus.x'='genus'))
hostTraits=subset(hostTraits, select=-c(order.y,family.y,genus.y))

##(7) merge transformed categorical variables and clean environment
hostTraits <- full_join(hostTraits, hostTraits_cat, by = c("order","family","genus"),keep=TRUE)
hostTraits <- rename(hostTraits,c('order.x'='order','family.x'='family','genus.x'='genus'))
hostTraits=subset(hostTraits, select=-c(order.y,family.y,genus.y))

##(8) clean environment
rm(hostTraits_binary,hostTraits_cat,hostTraits_continuous)

```

# Collapse hostTree to genus-level

```{r echo=F, message=T}

##(1) reformat
hostTree$tip.label[hostTree$tip.label=="_Anolis_carolinensis"] <- "Anolis_carolinensis"

##(2) create dataframe linking tip labels with their corresponding categories (genus and species)
tdata <- data.frame(matrix(NA,nrow=length(hostTree$tip.label),ncol=0))
tdata$genus=sapply(strsplit(hostTree$tip.label,'_'),function(x) paste(x[1],sep='_'))
tdata$species=hostTree$tip.label

##(3) collapse tree to genus level
hostTree=makeCollapsedTree(tree=hostTree,df=tdata[c('genus','species')])

##(4) clean environment
rm(tdata)

```

# Merge poxdata with hostTraits and hostTree

```{r echo=F, message=T}

## Before merging, we need to check for mismatched names between dataframes

##(1) are all poxdata genera in hostTree?
data$gtip <- data$gen
hostTree$gtip <- hostTree$tip.label
data$intree=ifelse(data$gtip%in%setdiff(data$gtip,hostTree$gtip),'missing','upham')

##(2) are all poxdata genera in hostTraits?
hostTraits$gtip <- hostTraits$genus
data$intraits <- ifelse(data$gtip%in%setdiff(data$gtip,hostTraits$gtip),'missing','traits')

##(3) create dataframe of just observations with mismatched names
fix=data[c('gtip','intree','intraits')]
fix=fix[fix$intree=='missing'|fix$intraits=='missing',]
fix=unique(fix)

##(4) identify homotypic synonyms or proxy species via IUCN (https://www.iucnredlist.org/) and NCBI (http://www.ncbi.nlm.nih.gov/taxonomy)
fix$traitname <- NA
fix$proxy <- NA
fix$proxy <- ifelse(fix$gtip=="Calassomys","Delomys",fix$proxy)                   #source: https://academic.oup.com/jmammal/article/95/2/201/860032
fix$traitname <- ifelse(fix$gtip=="Liomys","Heteromys",fix$traitname)             #source: https://www.iucnredlist.org/species/40768/22345036
fix$traitname <- ifelse(fix$gtip=="Oreonax","Lagothrix",fix$traitname)            #source: https://www.iucnredlist.org/species/39924/192307818
fix$traitname <- ifelse(fix$gtip=="Paralomys","Phyllotis",fix$traitname)          #source: https://www.iucnredlist.org/species/17226/22333354
fix$traitname <- ifelse(fix$gtip=="Pearsonomys","Geoxus",fix$traitname)           #source: https://www.iucnredlist.org/species/40768/22345036
fix$traitname <- ifelse(fix$gtip=="Pipanacoctomys","Tympanoctomys",fix$traitname) #source: https://www.iucnredlist.org/species/136557/78324400#taxonomy
fix$traitname <- ifelse(fix$gtip=="Pseudalopex","Lycalopex",fix$traitname)        #source: https://www.iucnredlist.org/species/6926/87695615
# hostTraits$genus[which(grepl('Tympanoctomys',hostTraits$genus))]

## merge revised names with poxdata
fix <- subset(fix, select=-c(intree,intraits))
data=merge(data,fix,by='gtip',all.x=T)

## if traitname is blank, NA
data$traitname=ifelse(data$traitname=='',NA,as.character(data$traitname))

## if traitname is NA, tip
data$traitname=ifelse(is.na(data$traitname),as.character(data$gtip),as.character(data$traitname))

## trait name includes proxy
data$traitname=ifelse(data$intraits=='missing' & is.na(data$traitname),as.character(data$proxy),
                      ifelse(data$intraits=='missing' & !is.na(data$traitname),as.character(data$traitname),
                             as.character(data$gtip)))
##(4) export missing names 
# write.csv(miss,'data/names/poxhost_name_mismatch.csv')
    # # to check whether species name exists in dataframe:
          # hostTree$tip.label[which(grepl('Pipanacoctomys',hostTree$tip.label))]
          hostTraits$genus[which(grepl('Tympanoctomys',hostTraits$genus))]
    # # to identify related species see IUCN (https://www.iucnredlist.org/) and NCBI (https://www.ncbi.nlm.nih.gov/taxonomy)

fix=read.csv('/Users/katietseng/Library/CloudStorage/OneDrive-WashingtonStateUniversity(email.wsu.edu)/Fernandez Lab/Projects (Active)/OPV Host Prediction/GitHub/PoxHost/EEID2022/data/names/poxhost_name_mismatch_KTedits.csv',header=TRUE)

```


# Merge viral accessory genes

```{r echo=F, message=T}

## (1) clean data
viralTraits <- opvgenes
viralTraits <- head(viralTraits, -2)          

## (2) rename column names
original_cols <- colnames(viralTraits)
colnames(viralTraits) <- paste("ag" ,original_cols,sep="_")
names(viralTraits)[1:2] <- c("virus","OPG")

## (3) to assess variation in viralTraits, create mode function
mode.prop <- function(x) {                
  ux <- unique(x[is.na(x)==FALSE])        # creates array of unique values
  tab <- tabulate(match(na.omit(x), ux))  # creates array of the frequency (number of times) a unique value appears in a column 
  max(tab)/length(x[is.na(x)==FALSE])     # max-frequency / number of elements in each column that are not NA
}

## (4) assess variation across columns (2 indicates columns)
vars=data.frame(apply(viralTraits,2,function(x) mode.prop(x)),
                apply(viralTraits,2,function(x) length(unique(x))))    # number of unique elements in each column
vars$variables=rownames(vars)
colnames(vars) <- c("var","uniq","column")

## trim
#vars <- vars[-c(1,2), ]

## drop variables with no variation
vars <- subset(vars,vars$var<1)

## visualize distribution of NA
png("/Users/katietseng/Downloads/virus_ag_variation.png", width=4,height=4,units="in",res=600)
ggplot(vars,
       aes(var))+
  geom_histogram(bins=50)+
  geom_vline(xintercept=0.70,linetype=2,size=0.5)+
  theme_bw()+
  theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank())+
  theme(axis.title.x=element_text(margin=margin(t=10,r=0,b=0,l=0)))+
  theme(axis.title.y=element_text(margin=margin(t=0,r=10,b=0,l=0)))+
  labs(y="frequency",
       x="trait coverage across viral species")+
  scale_x_continuous(labels = scales::percent)
dev.off()

## drop based on threshold
vars$keep=ifelse(vars$var>=0.7,"keep","cut")
keeps=vars[-which(vars$keep=="cut"),]$column #creates array of column names to cut and removes from df
keeps <- append("virus",keeps)
viralTraits=viralTraits[keeps]
rm(keeps,vars,mode.prop,original_cols,opvgenes)

# ## identify rows with duplicate values (i.e., hosts with identical presence/absence of accessory genes)
# which(duplicated(viralTraits[,-c(1)])| duplicated(viralTraits[,-c(1)], fromLast = TRUE))
# viralTraits$dup <- duplicated(viralTraits[,-c(1)])

## merge with poxdata
data <- merge(data,viralTraits,all.x=TRUE)

```


# Aggregate at genus level

```{r echo=F, message=T}
# pox_data <- virion
# taxa <- vertlife
# host_traits <- combine
# virus_traits <- opvgenes
# tree <- dryad
```

# Combine PubMed citations and evolutionary distinctiveness measure

```{r echo=F, message=T}
```

# Phylogenetic analysis

```{r echo=F, message=T}

# 02_phylofactor.R

```

# Inspect and format data for inclusion in the model

```{r echo=F, message=T}

```

# 

```{r echo=F, message=T}
```

# 

```{r echo=F, message=T}
```

# 

```{r echo=F, message=T}
```

# 