---
title: "Orthopoxvirus Link Prediction Model Code"
author: "Katie Tseng, Dan Becker, Colin Carlson, Pilar Fernandez, and Stephanie Seifert"
output:
  pdf_document:
  latex_engine: xelatex
  toc: yes
html_document:
  fig_height: 6
  fig_width: 6
  highlight: tango
  theme: journal
editor_options: 
  chunk_output_type: console
---

Introduction
============

The following code reproduces the analyses from *Viral genomic features predict orthopoxvirus reservoir hosts*, pertaining to the link prediction model. The code is subdivided into three parts (see Table of Contents below). To reproduce the analyses pertaining to the host prediction model, please see the markdown file *HostTraitModel_Code.Rmd* located in the [PoxHost repository][1].

Before running the following script, download the following data files to your working directory: 
- *~/data/linkpred_rawdata.RData*: the raw data file
- *~/data/OPVnew_nowwithVirus.xlsx*: table of OPV genome annotations extracted from NCBI with corresponding accessory gene data (binary matrix)
- *OPV_6Jan23.nexus*: the nexus file of viral tree data for visualizing clustering of OPV sequences on the phylogenetic tree
    - This file is only needed in part one for hierarchical cluster analysis (optional): *Dimension Reduction* {r dim_tree}

[1]: <https://github.com/viralemergence/PoxHost> "PoxHost"

Table of Contents
=================

1. Dimension Reduction {r dim}

2. Data Preparation {r prep}

3. BRT Model {r brt}

### *Before proceeding, we recommend setting knit options* 

```{r, knitr}

knitr::opts_chunk$set(eval=F)

```

1. Dimension Reduction
======================

In this section, we draw from our dataset of annotated Orthopoxvirus (OPV) genomes (n=197 unique sequences; see genome annotation pipeline in methods section for more information), which compose our known host-OPV associations for incorporation in the link prediction model. Each sequence is classified by its source or virus species and its host, as well as the presence (1) or absence (0) of a suite of OPV accessory genes. Because our data consists of over 981 OPV accessory genes, the goal of this section is to reduce the number of viral predictors for incorporation in the link prediction model, while maintaining maximal variance. Using principal components analysis (PCA), a method of dimension reduction, we distill the variables down to their most important features. We also explore which sequences have the most similar values (aka, how do they group?) in the principal components using k-means clustering. Because we are analyzing presence/absence data (binary variables), we use multiple correspondence analysis (MCA) to further validate our results. Though similar to PCA, MCA is used to analyze datasets with multiple categorical variables. 

### *Load required packages and set system*

```{r dim_load, message=FALSE, warning=FALSE}

# Libraries for preparing data for analysis
library(ape)
library(dplyr)
library(nlme)
library(tidyverse)
library(vroom) 
library(readxl)
library(ggplot2)

# Libraries for PCA (principal components analysis)
library(vegan)
library(factoextra) #fviz_eig
library(ggfortify)

# Libraries for MCA
library(FactoMineR)
library(dplyr)
library(factoextra) #fviz_eig

# Clean environment
rm(list=ls()) 
graphics.off()

# Set working directory
setwd("/Users/katietseng/Library/CloudStorage/OneDrive-WashingtonStateUniversity(email.wsu.edu)/Fernandez Lab/Projects (Active)/OPV Host Prediction/GitHub/PoxHost")

```

### *Prepare genomic data for dimension reduction*
We clean our sequence data and explore the variability in the data.
Figures: *histogram_var_genes.png*
Tables: *table_frequency_seqbyvirus.csv*

```{r dim_genes}

# Load genome annotations and trim
genes <- read_xlsx("data/OPVnew_nowwithVirus.xlsx", sheet="PoxHost")

# Rename variables and exclude unnecessary variables
genes <- plyr::rename(genes, c("Virus"="VirusSpecies","Host Genus"="HostGenus","Host Species"="HostSpecies"))
genes <- subset(genes, select=-c(HostSpecies))

# Correct sequence MT903347_1 - 'HostGenus' var lists Family name instead of Genus
genes$HostGenus <- ifelse(genes$HostGenus=="Gliridae","Graphiurus",genes$HostGenus)

# Add unique identifier
genes$rownames <- rownames(genes)
genes$Sequence <- paste(genes$Genome,genes$VirusSpecies,genes$HostGenus,sep="_",genes$rownames)
genes$rownames=NULL
genes <- genes %>% dplyr::select(Sequence, everything())

# View frequency of various virus species
prop_table <- subset(genes, select=-c(Sequence,Genome))
prop_table$Frequency = 1 
prop_table <- aggregate(Frequency ~ VirusSpecies + HostGenus, data=prop_table, FUN=sum)
prop_table <- prop_table[order(prop_table[,c("VirusSpecies")],prop_table[,c("HostGenus")]) ,]
prop_table$Perc <- prop_table$Frequency/sum(prop_table$Frequency)*100
print(prop_table)

# Save frequency table 
write.csv(prop_table, "figures/other/linkpred/table_frequency_seqbyvirus.csv")

# Create function (mode.prop) to assess variation in the presence/absence of OPV genes
mode.prop <- function(x) {                
  ux <- unique(x[is.na(x)==FALSE])        # creates array of unique values
  tab <- tabulate(match(na.omit(x), ux))  # creates array of the frequency (number of times) each unique value appears in the column 
  max(tab)/length(x[is.na(x)==FALSE])     # max-frequency / number of elements in each column that are not NA
}

# Assess variation across columns (2 indicates columns)
vars=data.frame(apply(genes,2,function(x) mode.prop(x)),
                apply(genes,2,function(x) length(unique(x)))) # number of unique elements in each column
vars$variables=rownames(vars)
colnames(vars) <- c("var","uniq","column")

# Trim
vars <- vars[-c(1:4), ]

# Any variables with no variation? If so drop
which(vars$var==1)
# vars <- subset(vars,vars$var<1)

# Visualize distribution of variation
gene_var <- ggplot(vars,
       aes(var))+
  geom_histogram(bins=50)+
  geom_vline(xintercept=0.70,linetype=2,linewidth=0.5)+
  theme_bw()+
  theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank())+
  theme(axis.title.x=element_text(margin=margin(t=10,r=0,b=0,l=0)))+
  theme(axis.title.y=element_text(margin=margin(t=0,r=10,b=0,l=0)))+
  labs(y="Frequency",
       x="Variance in the presence/absence of genes")+
  scale_x_continuous(labels=scales::percent)
gene_var

# Clean environment
rm(list=setdiff(ls(), c("genes")))

```

### *PCA of viral accessory genes*
Using PCA, can we distill the variables down to their most important features? Which genes contribute the most to each feature? Let's plot the loadings and also color them by functional role
Figure: *histogram_loadings.png*
Table: *table_pca_importance.csv*, *table_ranked_loadings.csv*

```{r dim_pca}

# Apply PCA using stats::prcomp 
pca <- prcomp(genes[,5:985])      #scaling/centering not appropriate
relvar <- pca$sdev^2 / sum(pca$sdev^2)
relvar_per <- round(relvar*100,1)

# View summary results
summary(pca)
View(pca$x) #sequence (individuals)
View(pca$rotation) #genes (variables)

# Get table of importance of components: eigenvalue (variance explained by each PC), SD, proportion of variance, and cumulative proportion
importance <- as.data.frame(t(summary(pca)$importance))
importance$Eigenvalue <- importance$`Standard deviation`^2
importance <- importance %>% dplyr::relocate(Eigenvalue) 
importance <- importance[c(1:12),]
write.csv(importance,"figures/other/linkpred/pca/table_pca_importance.csv")
rm(importance)

# Get table of loadings: rotation is the matrix of variable loadings where columns are eigenvectors
loadings <- as.data.frame(pca$rotation)
loadings <- loadings[,c(1:10)]
# loadings <- abs(loadings) #get absolute values; why are some loadings > |1|? Loading is the covariances/correlations b/w original vars and unit-scaled components)

### Create a df of ranked loadings and their gene names for each PC ###

  # For each PC, create a df of gene loadings named after the PC
  for(i in 1:ncol(loadings)){
    assign(colnames(loadings)[i], data.frame(loadings[,i]))
  }
  
  # Create a list of PC df names
  list <- colnames(loadings)    
  
  # Combine the dfs into a single list; drop PC dfs
  list_df = lapply(list, get)
  rm(list = ls()[grepl("PC", ls())])

  # To each list df, add corresponding gene name; sort in descending order
  for (i in 1:length(list)) {
    colnames(list_df[[i]]) <- "Loadings"
    list_df[[i]]$Gene <- rownames(loadings)
    list_df[[i]]=list_df[[i]][order(-list_df[[i]]$Loadings),]
  }
  
  # Create df of just ranked genes by first dropping loadings from the list
  genes_df <- list_df
  for(i in 1:length(list)) {
    genes_df[[i]]$Loadings=NULL
  }
  
  # Create an empty matrix with PCs as column names
  rank_genes <- data.frame(matrix(ncol=ncol(loadings), nrow=nrow(loadings)))
  colnames(rank_genes) <- colnames(loadings)
  
  # To each matrix column, add the ranked genes for each PC
  for(i in 1:length(list)) {
    rank_genes[,i] = genes_df[[i]]
  }
  
  # Set column names
  rank_genes <- setNames(rank_genes, paste0(names(loadings), '_', 'Gene'))
  
  # Create df of just ranked loadings by first dropping gene names from the list
  loadings_df <- list_df
  for(i in 1:length(list)) {
    loadings_df[[i]]$Gene=NULL
  }
  
  # Create an empty matrix with PCs as column names
  rank_loadings <- data.frame(matrix(ncol=ncol(loadings), nrow=nrow(loadings)))
  colnames(rank_loadings) <- colnames(loadings)
  
  # To each column of the matrix, add the ranked loadings for each PC
  for(i in 1:length(list)) {
    rank_loadings[,i] = loadings_df[[i]]
  }
  
  # Set column names
  rank_loadings <- setNames(rank_loadings, paste0(names(loadings), '_', 'Loadings'))
  
  # Combine dfs of ranked genes and loadings
  rank_PC <- cbind(rank_genes, rank_loadings)
  rank_PC <- rank_PC[,order(colnames(rank_PC))] 
  
  # Reorder columns
  rank_PC <- rank_PC %>% relocate(c("PC10_Gene","PC10_Loadings"), .after = last_col())
  
  # Save table of PCA Loadings Ranked
  write.csv(rank_PC, "figures/other/linkpred/pca/table_ranked_loadings.csv")

# Distribution of loadings for each PC (particularly PC1, PC3, PC4, PC9)
loadings <- as.data.frame(pca$rotation)
loadings <- loadings[,c(1:10)]
loadings_abs <- abs(loadings) #get absolute values

library(reshape2)
hist_loadings=ggplot(melt(loadings),aes(x=value)) + geom_histogram() + facet_wrap(~variable)
hist_loadings_abs=ggplot(melt(loadings_abs),aes(x=value)) + geom_histogram() + facet_wrap(~variable)
hist_loadings_abslim=ggplot(melt(loadings_abs),aes(x=value)) + geom_histogram() + xlim(0.05,0.20) + facet_wrap(~variable)

# Save figure
png("figures/other/linkpred/pca/histogram_loadings.png", width=4,height=4,units="in",res=600)
hist_loadings
dev.off()

# Save datafile of PC scores for link prediction
wgs_PCs <- genes[,1:4]
wgs_PCs <- cbind(wgs_PCs, pca$x[,1:10])
save(wgs_PCs, file='data/wgs_PCs.RData')

# Clean up environment
# rm(rank_genes, rank_loadings, genes_df, loadings, loadings_df, list_df, list)

# Clean environment
rm(list=setdiff(ls(), c("genes","pca","relvar","relvar_per", "rank_PC")))

```

### *PCA visualizations*
Which dimensions contribute the most?
Figures: *screeplot_var_pca.png*, *plot_cumvar_pca.png*, *plot_pca_scoresbyspecies.pdf*

```{r dim_pca_viz}

### Vizualize variance

# Screeplot of variance (eigenvalues) to show the decreasing rate at which variance is explained by additional PCs
png("figures/other/linkpred/pca/screeplot_var_pca.png", width=4,height=4,units="in",res=600)
screeplot(pca, type="lines", npcs=15, main="Scree plot of Eigenvalues for the first 15 PCs")
abline(h=1, col="red", lty=5)
legend("topright", legend=c("Eigenvalue = 1"), col=c("red"), lty=5, cex=1)
dev.off()
### suggests cutoff at PC12

# Screeplot of cumulative variance to show the % variance explained by additional PCs
screeplot_var <- barplot(relvar_per[1:10], xlab='PC', ylab='Percentage of explained variances', main='Screeplot of explained variances', names.arg=1:10, las=1, ylim=c(0,max(relvar_per)+10), col='gray')
text(screeplot_var, 0, y=relvar_per[1:10], label=relvar_per[1:10],cex=0.8, pos=3, col="red")

# Plot cumulative variance to show the proportion of variance explained with each add'l PC
cumpro <- cumsum(pca$sdev^2 / sum(pca$sdev^2))
png("figures/supplementary/s5fig_plot_cumvar_pca.png", width=4,height=4,units="in",res=600)
plot(cumpro[0:15], xlab = "Dimension", ylab = "Proportion of explained variance", main = "Cumulative variance plot")
abline(v = 10, col="blue", lty=5)
abline(h = 0.7, col="blue", lty=5)
legend("topleft", legend=c("Cut-off @ PC10"), col=c("blue"), lty=5, cex=1)
dev.off()

### Vizualize individuals/scores

# Style 1: Plot sequences on dim 1 and 2 (with single ellipses)
fviz_pca_ind(pca, geom.ind = "point", pointshape = 21, pointsize = 2, 
             col.ind = "black", addEllipses = TRUE, label = "var",
             col.var = "black", repel = TRUE,
             alpha.ind = 0.7) +
             ggtitle("PCA Plot of Sequences") +
             theme(plot.title = element_text(hjust = 0.5))

# Style 2: Plot sequences by virus species for dim 1 and 2
autoplot(pca, x=1, y=2, data = genes, colour = 'VirusSpecies') 

# Style 3: Plot sequences by virus species w/ viridis

  ## Color-blind friendly palette:
  library(viridis)
  cbPalette <- viridis(12, alpha = 1, begin = 0, end = 1, direction = 1, option = "H")

  ## Dim 1 and 2
  scores_dim12 <- fviz_pca_ind(pca, axes=c(1,2), title = "",
                  geom.ind = "point", pointshape = 21, pointsize = 2, 
                  fill.ind = genes$VirusSpecies, col.ind = "black", 
                  addEllipses = TRUE,label = "var", col.var = "black",repel = TRUE,
                  palette = cbPalette,
                  alpha.ind = 0.5) +
                  theme(legend.position="none")
  
  ## Dim 3 and 4
  scores_dim34 <- fviz_pca_ind(pca, axes=c(3,4), title = "",
                  geom.ind = "point", pointshape = 21, pointsize = 2, 
                  fill.ind = genes$VirusSpecies, col.ind = "black", 
                  addEllipses = TRUE,label = "var", col.var = "black",repel = TRUE,
                  palette = cbPalette,
                  alpha.ind = 0.5) +
                  theme(legend.position="none")
  
  ## Dim 5 and 6
  scores_dim56 <- fviz_pca_ind(pca, axes=c(5,6), title = "",
                  geom.ind = "point", pointshape = 21, pointsize = 2, 
                  fill.ind = genes$VirusSpecies, col.ind = "black", 
                  addEllipses = TRUE,label = "var", col.var = "black",repel = TRUE,
                  palette = cbPalette,
                  alpha.ind = 0.5) +
                  theme(legend.position="none")
  
  ## Dim 7 and 8
  scores_dim78 <- fviz_pca_ind(pca, axes=c(7,8), title = "",
                  geom.ind = "point", pointshape = 21, pointsize = 2, 
                  fill.ind = genes$VirusSpecies, col.ind = "black", 
                  addEllipses = TRUE,label = "var", col.var = "black",repel = TRUE,
                  palette = cbPalette,
                  alpha.ind = 0.5) +
                  theme(legend.position="none")

  ## Dim 9 and 10
  scores_dim910 <- fviz_pca_ind(pca, axes=c(9,10), title = "",
                  geom.ind = "point", pointshape = 21, pointsize = 2, 
                  fill.ind = genes$VirusSpecies, col.ind = "black", 
                  addEllipses = TRUE,label = "var", col.var = "black",repel = TRUE,
                  palette = cbPalette,
                  alpha.ind = 0.5) +
                  theme(legend.position="none")

  ## Get legend only
  legend <- cowplot::get_legend(fviz_pca_ind(pca, axes=c(9,10),
                  geom.ind = "point", pointshape = 21, pointsize = 2, 
                  fill.ind = genes$VirusSpecies, col.ind = "black", 
                  addEllipses = TRUE,label = "var", col.var = "black",repel = TRUE,
                  legend.title = "OPV species",
                  palette = cbPalette,
                  alpha.ind = 0.5)
                  
  ## Combine plots and save
  library(cowplot)
  plot_grid(scores_dim12, scores_dim34, scores_dim56, scores_dim78, scores_dim910, legend, ncol=3, nrow=2, scale=c(0.8), labels = c('A','B','C','D','E',''))  

### Vizualize variables/loadings (biplots)

# Style 1: Plot gene loadings only
 
  ## Dim 1 and 2
  fviz_pca_var(pca, 
               col.var = "contrib", # Color by contributions to the PC
               gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
               repel = TRUE,     # Avoid text overlapping
               # select.var=list(contrib=50),   # modify the number of loadings to include
               labelsize=1,
               label = c("ind", "ind.sup", "quali", "var", "quanti.sup"))
  dev.off()
    ### Here we see PC1 has large positive associations with a number of AGs like ADZ29556.1, SNB51281.1, and AGZ01283.1 that point in the same direction as PC1. PC2 has some moderately positive associations with AGs like BDQ10560.1
    ### QKE61192.1 - hypothetical protein [Vaccinia virus]
    ### QNP13375.1 - MPXV Viral membrane assembly proteins (VMAP) (Cop-A 30.5L)"

  ## Dim 3 and 4
  fviz_pca_var(pca, axes = c(3, 4),
               col.var = "contrib",
               gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
               repel = TRUE,
               # select.var=list(contrib=50),   # modify the number of loadings to include
               labelsize=1)
  dev.off()
  
# Style 2: Plot gene loadings by functional group
  
  ## Dim 1 and 2
  fviz_pca_var(pca, axes = c(1, 2),
               col.var = "contrib",
               gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
               repel = TRUE,
               # select.var=list(contrib=50),   # modify the number of loadings to include
               labelsize=1)
  dev.off()


# Biplot sequences and gene loadings
fviz_pca_biplot(pca, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                repel = TRUE,
                labelsize=1)
dev.off()

# Biplot top 20 influential scores and loadings
fviz_pca_biplot(pca, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, 
             labelsize=3,
             select.ind=list(contrib=20), select.var=list(contrib=20))
dev.off()

# Clean environment
rm(list=setdiff(ls(), c("genes","pca")))

```

### *PCA Hierarchical Cluster Analysis*
Which sequences have the most similar values? How do they group based on k-means clustering
Figures: *linegraph_optclusters_pca.png*, *clustree_pca.png*, *clustree_stability_pca.png*, *clustree_overpc1and2.png*, *cluster_dendogram.png*, *cluster_dendogram.png*, *plot_scores_pca_bycluster.png*
Tables: *table_scores_pca_bycluster.csv*

```{r dim_cluster}

### Determine Optimal Number of Clusters ###

library(clustree)

# Extract coordinates for individual sequences
ind.coord <- pca$x
rownames(ind.coord) <- genes$VirusSpecies
ind.coord <- ind.coord[,1:10]

###Elbow method
# Calculate/graph the sum of squares at each number of clusters. Assess how the slope changes from steep to shallow (an elbow) to determine the optimal number of clusters. How does increasing the number of the clusters contribute to separating the clusters in a meaningful way based on intra-cluster variance?

# Set seed and execute
set.seed(31)
png("figures/other/linkpred/pca/linegraph_optclusters_pca.png", width=4,height=4,units="in",res=600)
fviz_nbclust(ind.coord, kmeans, method = "wss", k.max = 20) + theme_minimal() 
# + ggtitle("the Elbow Method")
dev.off()

#### Clustree method
# How do samples change groupings as the number of clusters increases? Clustree shows which clusters are distinct and which are unstable. It doesn’t explicitly tell you which choice of optimal clusters is but it is useful for exploring possible choices.

# Execution of k-means with 1 to 10 clusters
tmp <- NULL
for (k in 1:10){
  tmp[k] <- kmeans(ind.coord, k, nstart = 30) #generates 30 initial configs & avg all centroid results 
}

# Convert to dataframe
df <- data.frame(tmp)

# Add a prefix to the column names of df
colnames(df) <- seq(1:10)
colnames(df) <- paste0("k",colnames(df))

# Combine clustering data w/ PC loading values
df <- bind_cols(as.data.frame(df), as.data.frame(ind.coord))

# Execute clustree
png("figures/other/linkpred/pca/clustree_pca.png", width=6,height=8,units="in",res=600)
clustree(df, prefix = "k")
dev.off()
### note: nodes with multiple incoming edges indicate that we over-clustered the data

# Execute clustree, but display stability index by colour as opposed to count
png("figures/other/linkpred/pca/clustree_stability_pca.png", width=6,height=8,units="in",res=600)
clustree(df, prefix = "k", node_colour = "sc3_stability")
dev.off()
### note: k=6 appears to be the most stable number of clusters

# Execute clustree_overlay, overlaying PC1 on PC2 to assess quality of the clustering
df_subset <- df %>% dplyr::select(1:10,11:12)
png("figures/other/linkpred/pca/clustree_overpc1and2.png", width=9, height=12, units="in", res=600)
clustree_overlay(df_subset, prefix = "k", x_value = "PC1", y_value = "PC2")
dev.off()


# Execute clustree overlaying PC vs. k (resolution dimension)
overlay_list <- clustree_overlay(df_subset, prefix = "k", x_value = "PC1",
                                 y_value = "PC2", plot_sides = TRUE)
overlay_list$x_side  #PC1
overlay_list$y_side  #PC2

### Conduct Hierarchical Cluster Analysis ###
# Each object is assigned its own cluster iteratively, at each stage joining the 2 most similar clusters (bottom-up dendogram until only one cluster is left)

# Extract coordinates for individual sequences
ind.coord <- pca$x
rownames(ind.coord) <- genes$VirusSpecies
ind.coord <- ind.coord[,1:10]

# Calculate distance matrix
pc.dist <- dist(ind.coord, method="euclidean")

# Execute HCA
clusters <- hclust(pc.dist)

# Plot dendogram
plot(clusters, cex=0.2)

# For k=6, add rectangular outline to each cluster in the dendrogram
png("figures/other/linkpred/pca/cluster_dendogram.png", width=8,height=5,units="in",res=600)
plot(clusters, cex=0.2)
rect.hclust(clusters, k = 6, border = 2:8) # add rectangle
dev.off() 

# For k=6, view count of each cluster
clusterCut <- cutree(clusters, k = 6)
table(clusterCut)

# Prop tables by virus species
mytable<-table(clusterCut, genes$VirusSpecies)
mytable2 <- data.frame(prop.table(mytable,2))
ggplot(mytable2, aes(x = Var2, y = Freq, fill = clusterCut)) +
  geom_col() +
  labs(fill='Cluster') +
  theme(axis.title.x = element_blank(), axis.text.x = element_text(angle=45,hjust=1)) +
  ggtitle("Distribution by Virus Species")

# Re-run PCA to color by cluster  
  #add cluster to original db
  genes1<-data.frame(cbind(genes,clusterCut))
  genes1$clusterCut <- as.factor(genes1$clusterCut)

# Run PCA as before, but now grouping by cluster
pca_bycluster <- prcomp(genes1[,5:985])   

# color-blind friendly palette:
cbPalette1 <- c("#009292","#ff6db6","#ffb6db","#490092","#6db6ff","#db6d00")
cbPalette2 <- c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00", "#CC79A7")

# Plot of sequences by cluster
png("figures/other/linkpred/pca/plot_scores_pca_bycluster.png", width=8,height=5,units="in",res=600)
fviz_pca_ind(pca_bycluster, geom.ind = "point", pointshape = 21, 
             pointsize = 2, 
             fill.ind = genes1$clusterCut, 
             col.ind = "black", 
             addEllipses = TRUE,
             label = "var",
             col.var = "black",
             repel = TRUE,
             legend.title = "Cluster",
             palette = cbPalette1,
             alpha.ind = 0.5)
dev.off()
  # + ggtitle("PCA Plot of Sequences by Cluster")

# Create a table of sequences by cluster
db_cluster <- dplyr::select(genes1, Genome, VirusSpecies, HostGenus,clusterCut)
write.csv(db_cluster,"figures/other/linkpred/table_scores_pca_bycluster.csv", row.names = F)

# Are the MPXV sequences in cluster 3 the same as the outliers in PCA4?
print(db_cluster[(db_cluster$VirusSpecies=="Monkeypox virus" & db_cluster$clusterCut==3),])
# print(outliers[,1:4])

# Clean environment
rm(list=setdiff(ls(), c("genes","genes1","mytable2", "pca")))

```

### *Visualize clustering on viral phylogenetic tree*
Figures: *virustree_barplot_cluster_pca.png*

```{r dim_tree}

# Resources
#https://github.com/YuLab-SMU/ggtree/issues/295

# Load libraries
library(ggtree)
library(tidyverse)
library(clustree)
library(ape) 
library(phylogram)

# Load viral tree data
vtree <- ape::read.nexus("data/OPV_6Jan23.nexus")
print(vtree$tip.label)

# Reformat tip label by dropping everything before the underscore
vtree$tip.label=substring(vtree$tip.label, regexpr("_", vtree$tip.label) + 1, nchar(vtree$tip.label))

# Repeat this step for tip labels with two underscores
vtree$tip.label=substring(vtree$tip.label, regexpr("_", vtree$tip.label) + 1, nchar(vtree$tip.label))

# Drop single quote in tip label
vtree$tip.label=gsub("'","",vtree$tip.label)

# Rename tip label Abatino to Abatino macacapox
vtree$tip.label <- ifelse(vtree$tip.label=="Abatino", "Abatino macacapox", vtree$tip.label)

# Rename tip label Cetacean to Cetaceanpox
vtree$tip.label <- ifelse(vtree$tip.label=="Cetacean", "Cetaceanpox", vtree$tip.label)

# Add ' virus' to tip label
vtree$tip.label=paste0(vtree$tip.label," virus")

# Convert tree data from class "phylo" to "dendrogram"
dendrogram <- as.dendrogram(vtree)

# Copy dataframe of cluster data (freq table of virus species by cluster)
mytable3 = mytable2

# Rename/reformat variables
mytable3$tip.label = mytable2$Var2
mytable3$Cluster = as.factor(mytable2$clusterCut)
mytable3$freq = mytable2$Freq
mytable3 <- subset(mytable3, select=c(tip.label,Cluster,freq))

# Check that all virus species names in mytable3 are in vtree
mytable3$intree <- ifelse(mytable3$tip.label%in%setdiff(mytable3$tip.label, vtree$tip.label),'missing','tree')
which(mytable3$intree=="missing")
mytable3$intree=NULL

# Plot of tree with scale of substitution rate
p <- ggtree(vtree) + geom_tiplab(align=TRUE, size=0) +
     geom_treescale(x=0.1, y=12, width=0.2, label="substitution rate") +
     annotate(geom="text", x=0.2, y=12.3, label="0.2", size=4)
# p <- ggtree(dendrogram) + geom_tiplab(align=TRUE, size=0)

# Color-blind friendly palette:
cbPalette1 <- c("#009292","#ff6db6","#ffb6db","#490092","#6db6ff","#db6d00")
cbPalette2 <- c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00", "#CC79A7")

# Plot of tree combined with bar plot of clustering
p1 <- facet_plot(p, data = mytable3, geom = geom_bar, 
                 panel = "Barplot", colour = "black", 
                 mapping = aes(fill=Cluster,x=freq), stat = "identity", orientation="y") +
                 scale_fill_manual(values=cbPalette1)

# Plots of tree + clustering with x-axis scales
p2 <- p1 + theme_tree2()

# Plots of tree + clustering with x-axis scales, x-axis titles, and tiplabel names
p3 <- p2 + xlab(label = c("\n      Substitutions per site                                                     Relative Proportion of Clustering")) +  
      theme(axis.title.x = element_text(size = 12)) + 
      geom_tiplab(as_ylab=TRUE, size = 10)
p3

# Save
png("figures/other/linkpred/pca/virustree_barplot_cluster_pca.png", width=16,height=8,units="in",res=600)
p3
dev.off()

```

### *PCA2 (Alternative Analysis)*
What happens when we exclude accessory genes present in only one virus species?

```{r dim_pca2}

# Drop accessory genes that are present in only one virus species (all 0's except for one)
genes2 <- genes[c(1:4,4 + which(colSums(genes[-(1:4)])>1))]
### 985 variables to 686 variables

# Apply PCA using stats::prcomp 
pca2 <- prcomp(genes2[,5:686])

### Vizualize variance, scores and loadings

# Plot cumulative variance to show the proportion of variance explained with each add'l PC
cumpro <- cumsum(pca2$sdev^2 / sum(pca2$sdev^2))
plot(cumpro[0:15], xlab = "Dimension", ylab = "Proportion of explained variance", main = "Cumulative variance plot")
abline(v = 10, col="blue", lty=5)
abline(h = 0.7, col="blue", lty=5)
legend("topleft", legend=c("Cut-off @ PC10"), col=c("blue"), lty=5, cex=1)

# Biplot sequences and gene loadings
fviz_pca_biplot(pca2, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
               repel = TRUE) +
               ggtitle("Biplot of Sequences and Gene Loadings")

# Biplot top 20 influential scores and loadings
fviz_pca_biplot(pca2, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, select.ind=list(contrib=20), select.var=list(contrib=20), max.overlaps=Inf) +
             ggtitle("Biplot of Top 20 Contributing Sequences and Gene Loadings")

### Summary: Compared to the first PCA, there's an increase in the proportion of variance explained by the first 10 dimensions from 0.696 to 0.712. No noticeable difference in the spatial distribution of scores and vectors. 

```

### *PCA3 (Alternative Analysis)*
What happens when we drop duplicate observations within the same host-virus links (sequences with the same identical presence/absence of accessory genes as another sequence of the same host-virus link)?

```{r dim_pca3}

# Identify observations of the same host-virus links with identical presence/absence of accessory genes
genes3 <- genes
genes3$dup <- duplicated(genes3[,-c(1:2)])
table(genes3$dup)
### 42 dups

# Drop duplicate observations
genes3 <- genes3[genes3$dup==FALSE,]
genes3$dup=NULL
### 197 obs to 155 obs

# Apply PCA using stats::prcomp 
pca3 <- prcomp(genes3[,5:985])

### Vizualize variance, scores and loadings

# Plot cumulative variance to show the proportion of variance explained with each add'l PC
cumpro <- cumsum(pca3$sdev^2 / sum(pca3$sdev^2))
plot(cumpro[0:15], xlab = "Dimension", ylab = "Proportion of explained variance", main = "Cumulative variance plot")
abline(v = 10, col="blue", lty=5)
abline(h = 0.7, col="blue", lty=5)
legend("topleft", legend=c("Cut-off @ PC10"), col=c("blue"), lty=5, cex=1)

# Biplot sequences and gene loadings
fviz_pca_biplot(pca3, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
               repel = TRUE) +
               ggtitle("Biplot of Sequences and Gene Loadings")

# Biplot top 20 influential scores and loadings
fviz_pca_biplot(pca3, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, select.ind=list(contrib=20), select.var=list(contrib=20), max.overlaps=Inf) +
             ggtitle("Biplot of Top 20 Contributing Sequences and Gene Loadings")

### Summary: Compared to the first PCA, there is a decrease in the proportion of variance explained by the first ten PCs from 0.696 to 0.680. As in previous PCAs, biplot of top 20 contributing sequences and gene loadings show the most influential sequences in the third quadrant. 

```

### *PCA4 (Alternative Analysis )*
What happens if we exclude the potential outliers from PCA, and then predict their scores and loadings?

```{r dim_pca4}

# Create df excluding outliers identified in PCA3
genes4 <- genes[!grepl("MT724769_1|MN346703_1|MT724770_1|DQ011155_1", genes$Genome),]

# Create df of outliers
outliers <- genes[grepl("MT724769_1|MN346703_1|MT724770_1|DQ011155_1", genes$Genome),]

# Apply PCA using stats::prcomp
pca4 <- prcomp(genes4[,5:985])
relvar <- pca4$sdev^2 / sum(pca4$sdev^2)
relvar_per <- round(relvar*100,1)

# Prediction of PCs for outliers
pred <- predict(pca4, newdata=outliers)
pca4_pred <- pca4
pca4_pred$x <- rbind(pca4_pred$x, pred)

### Vizualize scores and loadings

# Plot of individuals by virus species w/ outliers in shaded bullets
COLOR <- c(1:length(unique(genes$VirusSpecies)))
pc <- c(1,2)
plot(pca4$x[,pc], cex=1, col=COLOR,
     xlab=paste0("PC 1", "(", relvar_per[pc[1]], "%)"), 
     ylab=paste0("PC 2", "(", relvar_per[pc[2]], "%)"))
points(pred[,pc], pch=16) + abline(h = 0, v=0, lty = 2) +
title("Plot of Sequences and Outliers") + theme(plot.title = element_text(hjust = 0.5))

# Biplot of individuals and variables
fviz_pca_biplot(pca4_pred, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE) +
             ggtitle("Biplot of Sequences and Gene Loadings")

# Biplot of top 20 contributing individuals and variables
fviz_pca_biplot(pca4_pred, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, select.ind=list(contrib=20), select.var=list(contrib=20)) +
             ggtitle("Biplot of Sequences and Gene Loadings")

# Clean environment
rm(list=setdiff(ls(), c("genes", "pca","outliers")))

### Summary: Predicted scores of outliers cluster in the fourth quadrant with other sequences. As in previous PCAs, biplot of top 20 contributing sequences and gene loadings show the most influential sequences in the third quadrant.

```

### *MCA of viral accessory genes*
Multiple Correspondence Analysis (MCA) for dimension reduction of categorical variables. MCA is different from PCA in that it's primarily used for categorical data, which in this example are the presence absence of the accessory genes. MCA also uses a dissimilarity measure as opposed to a covariance or correlation matrix to calculate the dissimilarities between categories, which are measured by the singular value decomposition (SVD) of the data matrix (as opposed to eigenvalues or eigenvectors).

```{r dim_mca}

# Subset data and reformat gene variables as factor
genes_cat <- subset(genes,select=-c(Genome,VirusSpecies,HostGenus))
genes_cat[] <- lapply(genes_cat, as.character)
rownames <- genes$Sequence
genes_cat[,-1] <- lapply(genes_cat[,-1], factor)
genes_cat$Sequence=NULL
rownames(genes_cat) <- rownames
#str(genes_cat)

# Apply MCA using FactoMineR::MCA
mca = MCA(genes_cat, graph = FALSE)
# pca_relvar <- pca$sdev^2 / sum(pca$sdev^2)
# pca_relvar_per <- round(pca_relvar*100,1)

# List and summarize MCA results
print(mca)
# summary(mca)
head(mca$ind$coord) #sequence (individuals)
head(mca$var$coord) #genes (variables)

# Screeplot - Variance (Eigenvalues)
#mca$eig
fviz_eig(mca, addlabels = TRUE, ylim = c(0, 25))

# Plots of individuals
fviz_mca_ind(mca, repel=TRUE)

#(6) Plots of MCA variables 1 and 2
fviz_mca_var(mca, repel = TRUE) ##

# Biplot
fviz_mca_biplot(mca, repel = TRUE)
fviz_mca_biplot(mca, repel = FALSE, select.ind=list(contrib=20), select.var=list(contrib=20))

# Clean environment
rm(list=ls())

```


2. Data Preparation
===================

### *Load packages and data for merging: i.e., Host-pox/viral traits (PC), taxonomy data, host traits, and host tree*

```{r prep_raw}

# Load treespace
## treespace dependencies include XQuartz v2.7.11 (https://www.xquartz.org/releases/XQuartz-2.7.11.html) and 'rgl' (https://stackoverflow.com/a/66127391/2554330)
## recommend installing and loading rgl including 'options(rgl.useNULL=TRUE)' below before loading treespace
library(rgl) # > install.packages("rgl"); > options(rgl.useNULL=TRUE)
library(treespace) 
library(dplyr)

# Clean environment
rm(list=ls()) 
graphics.off()

# Set working directory
setwd("/Users/katietseng/Library/CloudStorage/OneDrive-WashingtonStateUniversity(email.wsu.edu)/Fernandez Lab/Projects (Active)/OPV Host Prediction/GitHub/PoxHost")

# Load raw data
load("data/linkpred_rawdata.RData")

# Load genomic sequence data
load("data/wgs_PCs.RData")

# Pox data: host-OPV interactions detected via PCR/isolation from Virion database *note additional interactions from pc_genes below
##virion <- vroom('https://github.com/viralemergence/virion/blob/main/Virion/Virion.csv.gz')
poxdata <- virion %>% filter(VirusGenus == "orthopoxvirus" & (DetectionMethod %in% c("PCR/Sequencing","Isolation/Observation"))) 

# Viral traits data (& Host-pox links): host-OPV linked interactions & principal components of OPV genes (viral traits) extracted from genome sequence data 
wgsdata <- wgs_PCs

# Taxa: mammal species taxonomy from vertlife
##vertlife <- read.csv(url('https://data.vertlife.org/mammaltree/taxonomy_mamPhy_5911species.csv'))
taxa <- vertlife

# Host traits: mammal traits from the COMBINE database <https://doi.org/10.1002/ecy.3344>
##path: ecy3344-sup-0001-datas1.zip > COMBINE_archives > trait_data_imputed.csv)
hostTraits <- combine

# Host tree: mammal phylogeny tree from Dryad, <https://doi.org/10.5061/dryad.tb03d03>
##path: Data_S8_finalFigureFiles > _DATA > MamPhy_fullPosterior_BDvr_Completed_5911sp_topoCons_NDexp_MCC_v2_target.tre)
hostTree <- dryad

# Clean environment
rm(virion, wgs_PCs, vertlife, dryad, combine)

```

### *Merge host-OPV interactions from Virion with host-OPV interactions from genome annotations (Stephanie)*

```{r prep_poxdata}

### Clean up Virion data ###

# Exclude if host genus or virus is NA
poxdata <- poxdata[!is.na(poxdata$HostGenus),]
poxdata <- poxdata[!is.na(poxdata$Virus),]

# Exclude variola (smallpox) virus
poxdata <- poxdata[!(poxdata$Virus=="variola virus"),]

# Reclassify feline poxvirus as cowpox virus
#In this model, we keep vaccinia virus!

# Aggregate data at the host genus level
poxdata$link <- 1
poxdata <- aggregate(link~HostGenus+Virus, data=poxdata, mean)
poxdata$link=NULL

# Rename and reformat variables
poxdata <- plyr::rename(poxdata,c('HostGenus'='genus','Virus'='virus'))
poxdata$genus <- str_to_title(poxdata$genus)

# Add datasource variable
poxdata$virion <- 1

### Clean up genomic data ###

# Rename and reformat variables
wgsdata <- plyr::rename(wgsdata,c('HostGenus'='genus','Sequence'='sequence'))
wgsdata$virus <- tolower(wgsdata$VirusSpecies)
wgsdata$VirusSpecies=NULL

# Drop genome variable
wgsdata$Genome=NULL

# Add datasource variable
wgsdata$wgs <- 1

### Merge ###

# Are their virus in wgsdata that are not in poxdata and vice versa?
wgsdata$virus[!wgsdata$virus %in% poxdata$virus]
poxdata$virus[!poxdata$virus %in% wgsdata$virus]

# Correct 'cetaceanpox virus' to 'cetacean pox virus 1' in wgsdata
wgsdata$virus <- ifelse(wgsdata$virus=="cetaceanpox virus", "cetacean poxvirus 1", wgsdata$virus)

# Are their genera in wgsdata that are not in poxdata and vice versa?
wgsdata$genus[!wgsdata$genus %in% poxdata$genus]
poxdata$genus[!poxdata$genus %in% wgsdata$genus]

# Before merging, get count of unique host-virus associations in each dataset
poxdata_unique <- unique(poxdata[ , c("genus", "virus")])
length(unique(poxdata_unique$genus))
length(unique(poxdata_unique$virus))
wgsdata_unique <- unique(wgsdata[ , c("genus", "virus")])
length(unique(wgsdata_unique$genus))
length(unique(wgsdata_unique$virus))

# Merge poxdata w/ wgsdata
poxdata <- merge(poxdata,wgsdata,by=c('virus','genus'),all=TRUE)

# Create data source variable
poxdata$source <- ifelse(poxdata$wgs==1 & is.na(poxdata$virion),"wgs", ifelse(poxdata$virion==1 & is.na(poxdata$wgs),"virion","both"))

# Replace sequence==NA with "NA", as this var will be our unique identifier for WGS links
poxdata$sequence <- ifelse(is.na(poxdata$sequence),"NA",poxdata$sequence)

# Get count of unique host-virus associations in each dataset
poxdata_unique <- unique(poxdata[ , c("genus", "virus")])
length(unique(poxdata_unique$genus))
length(unique(poxdata_unique$virus))

# Remove virion and wgs vars; clean environment
poxdata$virion=NULL
poxdata$wgs=NULL
rm(wgsdata, wgsdata_unique, poxdata_unique)

```

### *Merge poxdata with broader mammal taxa to create pseudoabsences*

```{r prep_pseudo}

# Drop duplicate genera in taxa
gtaxa <- taxa[!duplicated(taxa$gen),]
gtaxa <- gtaxa[c('gen','fam','ord')]
gtaxa <- plyr::rename(gtaxa, c('gen'='genus'))

# Check for mismatched genus names between poxdata and taxa before merging poxdata with taxa
poxdata$genus[!poxdata$genus %in% gtaxa$genus]
poxdata <- merge(gtaxa,poxdata,by='genus',all.x=TRUE)

# To keep only genera from orders in which positive associations exist, first subset known host-virus associations
keep <- subset(poxdata, !is.na(poxdata$source))

# Next, create a new variable <keep> in poxdata denoting observations with the same host taxonomic order as that of known associations
poxdata$keep <- ifelse(poxdata$ord %in% keep$ord,TRUE,FALSE)

# View/evaluate which taxonomics orders will be kept and which orders will be discarded
uniq <- unique(poxdata[c("ord","keep")])

# Keep only observations with the same host taxonomic order as that of known associations
poxdata <- subset(poxdata,keep==TRUE)
poxdata$keep=NULL

# Create binary variable for known host-OPV associations/links
poxdata$link=ifelse(is.na(poxdata$source),0,1)

# Reorder variables
poxdata <- poxdata %>% select(link, virus, genus, fam, ord, source, everything())

# Clean environment
rm(taxa,gtaxa,keep,uniq)

```

### *Aggregate hostTraits to genus-level*

```{r prep_traits}

# Observe variable names
colnames(hostTraits)

# To aggregate continuous/integer variables, use the median as the summary measure
hostTraits_continuous=aggregate(cbind(adult_mass_g,brain_mass_g,adult_body_length_mm,adult_forearm_length_mm,
                                   max_longevity_d,maturity_d,female_maturity_d,male_maturity_d,
                                   age_first_reproduction_d,gestation_length_d,teat_number_n,
                                   litter_size_n,litters_per_year_n,interbirth_interval_d,
                                   neonate_mass_g,weaning_age_d,weaning_mass_g,generation_length_d,
                                   dispersal_km,density_n_km2,home_range_km2,social_group_n,
                                   dphy_invertebrate,dphy_vertebrate,dphy_plant,
                                   det_inv,det_vend,det_vect,det_vfish,det_vunk,det_scav,det_fruit,det_nect,det_seed,det_plantother,det_diet_breadth_n,
                                   upper_elevation_m,lower_elevation_m,altitude_breadth_m,habitat_breadth_n) 
                             ~ order+family+genus, data=hostTraits, FUN=median, na.action=na.pass, na.rm=TRUE)
##'na.action=na.pass, na.rm=TRUE' is specified such that if species w/in a genus has a combination of real values & NAs, the median of real values will be returned (as opposed to omitting the genus or returning NA)

# To aggregate binary variables, use the mean as the summary measure
hostTraits$fossoriality[hostTraits$fossoriality==2]<-0  #recode 0/1: 0=above ground, 1=fossorial dwelling
hostTraits_binary=aggregate(cbind(hibernation_torpor,fossoriality,freshwater,marine,terrestrial_non.volant,terrestrial_volant,island_dwelling,disected_by_mountains,glaciation) ~ order+family+genus, data=hostTraits, FUN=mean, na.action=na.pass, na.rm=TRUE)
hostTraits_binary_temp=aggregate(cbind(hibernation_torpor,fossoriality,freshwater,marine,terrestrial_non.volant,terrestrial_volant,island_dwelling,disected_by_mountains,glaciation) ~ order+family+genus, data=hostTraits, FUN=mean, na.action=na.pass, na.rm=TRUE)

# To aggregate categorical variables, first transform the variables to binary
hostTraits_cat <- hostTraits
hostTraits_cat$trophic_herbivores <- ifelse(hostTraits_cat$trophic_level==1,1,0)
hostTraits_cat$trophic_omnivores <- ifelse(hostTraits_cat$trophic_level==2,1,0)
hostTraits_cat$trophic_carnivores <- ifelse(hostTraits_cat$trophic_level==3,1,0)
hostTraits_cat$activity_nocturnal <- ifelse(hostTraits_cat$activity_cycle==1,1,0)
hostTraits_cat$activity_crepuscular <- ifelse(hostTraits_cat$activity_cycle==2,1,0) #nocturnal/crepuscular, cathemeral, crepuscular or diurnal/crepuscular
hostTraits_cat$activity_diurnal <- ifelse(hostTraits_cat$activity_cycle==3,1,0)
hostTraits_cat$forager_marine <- ifelse(hostTraits_cat$foraging_stratum=="M",1,0)
hostTraits_cat$forager_ground <- ifelse(hostTraits_cat$foraging_stratum=="G",1,0) 
hostTraits_cat$forager_scansorial <- ifelse(hostTraits_cat$foraging_stratum=="S",1,0)
hostTraits_cat$forager_arboreal <- ifelse(hostTraits_cat$foraging_stratum=="Ar",1,0)
hostTraits_cat$forager_aerial <- ifelse(hostTraits_cat$foraging_stratum=="A",1,0)
hostTraits_cat$island_end_marine <- ifelse(hostTraits_cat$island_endemicity=="Exclusively marine",1,0)
hostTraits_cat$island_end_mainland <- ifelse(hostTraits_cat$island_endemicity=="Occurs on mainland",1,0)
hostTraits_cat$island_end_lgbridge <- ifelse(hostTraits_cat$island_endemicity=="Occurs on large land bridge islands",1,0)
##hostTraits_cat$island_end_smbridge <- ifelse(hostTraits_cat$island_endemicity=="Occurs on small land bridge islands",1,0)
hostTraits_cat$island_end_isolated <- ifelse(hostTraits_cat$island_endemicity=="Occurs only on isolated islands",1,0)
hostTraits_cat$biogeo_afrotropical <- ifelse(grepl("Afrotropical",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_antarctic <- ifelse(grepl("Antarctic",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_australasian <- ifelse(grepl("Australasian",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_indomalayan <- ifelse(grepl("Indomalayan",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_nearctic <- ifelse(grepl("Nearctic",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_neotropical <- ifelse(grepl("Neotropical",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_oceanian <- ifelse(grepl("Oceanian",hostTraits_cat$biogeographical_realm),1,0)
hostTraits_cat$biogeo_palearctic <- ifelse(grepl("Palearctic",hostTraits_cat$biogeographical_realm),1,0)

# To aggregate transformed categorical-to-binary variables, use the mean as the summary measure
hostTraits_cat=aggregate(cbind(trophic_herbivores,trophic_omnivores,trophic_carnivores,
                            activity_nocturnal,activity_crepuscular,activity_diurnal,
                            forager_marine,forager_ground,forager_scansorial,forager_arboreal,forager_aerial,
                            island_end_marine,island_end_mainland,island_end_lgbridge,island_end_isolated,
                            biogeo_afrotropical,biogeo_antarctic,biogeo_australasian,biogeo_indomalayan,biogeo_nearctic,biogeo_neotropical,biogeo_oceanian,biogeo_palearctic)
                       ~ order+family+genus, data=hostTraits_cat, FUN=mean, na.action=na.pass, na.rm=TRUE)

# Merge continuous variables with binary variables and simplify dataframe
hostTraits <- full_join(hostTraits_continuous, hostTraits_binary, by = c("order","family","genus"),keep=TRUE)
hostTraits <- plyr::rename(hostTraits,c('order.x'='order','family.x'='family','genus.x'='genus'))
hostTraits=subset(hostTraits, select=-c(order.y,family.y,genus.y))

# Merge transformed categorical variables and simplify dataframe
hostTraits <- full_join(hostTraits, hostTraits_cat, by = c("order","family","genus"),keep=TRUE)
hostTraits <- plyr::rename(hostTraits,c('order.x'='order','family.x'='family','genus.x'='genus'))
hostTraits <- subset(hostTraits, select=-c(order.y,family.y,genus.y))

# Clean environment
rm(hostTraits_binary,hostTraits_cat,hostTraits_continuous)

```

### *Trim hostTree to genus-level*

```{r prep_tree}

# Reformat
hostTree$tip.label[hostTree$tip.label=="_Anolis_carolinensis"] <- "Anolis_carolinensis"

# Create dataframe linking tip labels with their corresponding categories (genus and species)
tdata <- data.frame(matrix(NA,nrow=length(hostTree$tip.label),ncol=0))
tdata$genus <- sapply(strsplit(hostTree$tip.label,'_'),function(x) paste(x[1],sep='_'))
tdata$species <- hostTree$tip.label

# Collapse tree to genus level
hostTree <- makeCollapsedTree(tree=hostTree,df=tdata[c('genus','species')])

# Clean environment
rm(tdata)

```

### *Check for mismatched genera names in poxdata, hostTraits and hostTree*

```{r prep_names}

# Check if all poxdata genera are in hostTree
poxdata$gtip <- poxdata$genus
hostTree$gtip <- hostTree$tip.label
poxdata$intree <- ifelse(poxdata$gtip%in%setdiff(poxdata$gtip,hostTree$gtip),'missing','upham')

# Check if all poxdata genera are in hostTraits
hostTraits$gtip <- hostTraits$genus
poxdata$intraits <- ifelse(poxdata$gtip%in%setdiff(poxdata$gtip,hostTraits$gtip),'missing','traits')

# Create a dataframe of just the observations with mismatched names
fix <- poxdata[c('gtip','intree','intraits')]
fix <- fix[fix$intree=='missing'|fix$intraits=='missing',]
fix <- unique(fix)

# For those with mismatched names, identify homotypic synonyms or proxy species via IUCN (https://www.iucnredlist.org/) and NCBI (http://www.ncbi.nlm.nih.gov/taxonomy)
fix$treename <- NA
fix$traitname <- NA
fix$proxy <- NA
fix$proxy <- ifelse(fix$gtip=="Calassomys","Delomys",fix$proxy)
  ##source: https://academic.oup.com/jmammal/article/95/2/201/860032
fix$traitname <- ifelse(fix$gtip=="Liomys","Heteromys",fix$traitname)
  ##source: https://www.iucnredlist.org/species/40768/22345036
fix$traitname <- ifelse(fix$gtip=="Oreonax","Lagothrix",fix$traitname)
  ##source: https://www.iucnredlist.org/species/39924/192307818
fix$traitname <- ifelse(fix$gtip=="Paralomys","Phyllotis",fix$traitname)
  ##source: https://www.iucnredlist.org/species/17226/22333354
fix$traitname <- ifelse(fix$gtip=="Pearsonomys","Geoxus",fix$traitname)
  ##source: https://www.iucnredlist.org/species/40768/22345036
fix$traitname <- ifelse(fix$gtip=="Pipanacoctomys","Tympanoctomys",fix$traitname)
  ##source: https://www.iucnredlist.org/species/136557/78324400#taxonomy
fix$traitname <- ifelse(fix$gtip=="Pseudalopex","Lycalopex",fix$traitname)
  ##source: https://www.iucnredlist.org/species/6926/87695615
## hostTraits$genus[which(grepl('Tympanoctomys',hostTraits$genus))]

# Merge revised names with poxdata
fix <- subset(fix, select=-c(intree,intraits))
poxdata <- merge(poxdata,fix,by='gtip',all.x=T)

# If 'treename' is missing, first relabel as NA, then relabel with 'gtip'
poxdata$treename <- ifelse(poxdata$treename=='',NA,as.character(poxdata$treename))
poxdata$treename <- ifelse(is.na(poxdata$treename),as.character(poxdata$gtip),as.character(poxdata$treename))

# If 'traitname' is missing, first relabel as NA; If 'traitname' is NA and missing in 'intraits', then relabel with 'proxy'; If 'traitname' is not NA and missing in 'intraits', then relabel with 'traitname'
poxdata$traitname <- ifelse(poxdata$traitname=='',NA,as.character(poxdata$traitname))
poxdata$traitname <- ifelse(poxdata$intraits=='missing' & is.na(poxdata$traitname),as.character(poxdata$proxy),
                      ifelse(poxdata$intraits=='missing' & !is.na(poxdata$traitname),as.character(poxdata$traitname),
                             as.character(poxdata$gtip)))

# Simplify and clean environment
poxdata <- subset(poxdata, select=-c(intree,intraits,proxy))
rm(fix)

```

### *Merge poxdata with hostTraits and trim hostTree to mirror poxdata*

```{r prep_merge}

# Merge traits with poxdata
hostTraits$traitname <- hostTraits$gtip
poxdata <- merge(poxdata,hostTraits,by=c('traitname'),all.x=T)

# Clean up poxdata
poxdata <- plyr::rename(poxdata,c('gtip.x'='gtip','genus.x'='genus'))
poxdata <- subset(poxdata,select=-c(order, family, genus.y, gtip.y))

# Trim hostTree (remove species tip) to mirror poxdata
hostTree <- keep.tip(hostTree,hostTree$tip.label[hostTree$tip.label%in%poxdata$treename])
hostTree$gtip <- NULL
hostTree=makeLabel(hostTree)

# Clean environment
rm(hostTraits)

```

### *Add PubMed citations and evolutionary distinctiveness measure*

```{r prep_cited}

# Load library for PubMed citations
library(easyPubMed)

# Create function to count citations
counter=function(name){
  as.numeric(as.character(get_pubmed_ids(gsub('_','-',name))$Count))
}
citations=c()

# Extract unique genera from poxdata
treename <- unique(poxdata$treename)

# Apply counter function while looping through treenames
for(i in 1:length(treename)) {
  citations[i]=counter(treename[i])
  print(i)
}

# Compile citation numbers
cites <- data.frame(treename=treename,cites=citations)

# Merge cites with poxdata
poxdata <- merge(poxdata,cites,by='treename')

# Load library for evolutionary distinctiveness (ed) measure
library(picante)  #before loading picante, make sure latest version of nlme package is loaded
ed <- evol.distinct(hostTree,type='equal.splits') #calculates ed measures for a suite of species by equal splits and fair proportions; returns species score

# Rename variables in ed
ed <- plyr::rename(ed,c('Species'='treename','w'='ed_equal'))

# Merge ed with poxdata
poxdata <- merge(poxdata,ed,by='treename')

# Clean environment
rm(cites,ed,citations,i,treename,counter)

## Other potential viral traits to consider for future studies include: viral genome length, viral richness (number of virus detected in each genera), and host range (number of hosts from which each virus was detected)

```

### *Add all possible host-OPV combinations (i.e., pseudoabsences) for link prediction model and merge with PCA (viral genomic) data*

```{r prep_link}

# Create a dataframe of all possible host-OPV combinations (for mammal genera that exist in orders w/ known OPV predictions) by identifying all unique virus species and host genus of known associations and using the {expand.grid} function
uniq_virus <- unique(poxdata$virus[!is.na(poxdata$virus)])
uniq_genus <- unique(poxdata$genus[!is.na(poxdata$genus)])
combinations <- expand.grid(uniq_virus, uniq_genus)
combinations <- plyr::rename(combinations,c('Var1'='virus','Var2'='genus'))

# Create two dataframes: one subsetting host-OPV interaction/link data from poxdata (excluding host trait data), and the other subsetting unique host trait data (i.e., all other variables) which we will merge back in later
linkdata <- poxdata[,grepl("link|virus|genus|source|sequence|PC", names(poxdata))]
hostTraits <- poxdata[,!grepl("link|virus|source|sequence|PC", names(poxdata))] # include genus b/c we will merge on it later
hostTraits <- hostTraits[!duplicated(hostTraits$genus),]

# Merge linkdata with all possible combinations (drop observations if virus NA)
linkdata <- merge(linkdata, combinations, by=c("virus","genus"),all=TRUE)
linkdata <- linkdata[!is.na(linkdata$virus),]

# Merge linkdata with hostTraits
linkdata <- merge(linkdata, hostTraits, by=c("genus"))

# Reorder variables
linkdata <- linkdata %>% select(source, sequence, link, virus, genus, fam, ord, gtip, traitname, treename, PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10, everything())

# Reclassify NAs as pseudo-absences for viral detection
linkdata$link=ifelse(is.na(linkdata$link),0,linkdata$link)

### Next, we need to impute values of PC variables where data exist for that virus.

# First let's subset virus and PC variables, and drop rows w/ NA
pc <- linkdata[,grepl("virus|PC", names(linkdata))]
pc <- pc[!is.na(pc$PC1),]

# Let's take a quick look at which virus do not have PC data
setdiff(unique(linkdata$virus),unique(pc$virus))

# Next, we obtain the median value of each PC variable for each virus using aggregate function (dot notation), and rename PC variables
pc_median=aggregate(. ~virus, data=pc, FUN=median)
colnames(pc_median)[c(-1)] <- paste(colnames(pc_median)[c(-1)], "_med", sep="")

# Merge PC median variables w/ poxdata
linkdata <- merge(linkdata,pc_median,by=c("virus"),all.x=TRUE)

# Get PC colnames and replace NA values of PC vars w/ median PC values using for-loop
pc_names <- colnames(pc)[c(-1)]
for(i in pc_names) {
  linkdata[,i] <- ifelse(is.na(linkdata[,i]),linkdata[,paste(i,"_med",sep="")],linkdata[,i])
}

# Drop median PC variables
pc_med_names <- colnames(pc_median)[c(-1)]
for(i in pc_med_names) {
  linkdata[,i]=NULL
}

# Replace NA values for 'source' and 'sequence' with "NA"
linkdata$source <- ifelse(is.na(linkdata$source),"NA",linkdata$source)
linkdata$sequence <- ifelse(is.na(linkdata$sequence),"NA",linkdata$sequence)

# Clean environment
rm(combinations, hostTraits, uniq_virus, uniq_genus, pc, pc_median, pc_names, pc_med_names, i, poxdata)

```

### *Save cleaned data*

```{r prep_save}

#(1) Reorder variables
linkdata <- linkdata %>% select(source, sequence, link, virus, genus, fam, ord, gtip, traitname, treename, cites, ed_equal, everything())

#(2) Save dataframes for analysis **poxdata_temp.RData is for practice analysis**
save(linkdata, hostTree, file='data/linkpred_cleandata.RData')

```


3. BRT Model
============
This chapter builds boosted regression tree models to predict host-virus links. The following section is coded to run on your local computer and will likely take ~72+ hours (with parallel processing on 5 cores), outputting an RData file size of 2GB. For tips on running the code on an available HPC node, see 'Tseng2022/HPC Example' on the PoxHost GitHub repository: https://github.com/viralemergence/PoxHost/tree/0a1effef83dbd5f6f3d88c6d0c15c563eb499452/Tseng2022/HPC%20Example_01Jun2023.

### *Load required packages and set system*

```{r brt_load}

#(1) Libraries for BRT model
library(gbm)
library(fastDummies)
library(rsample)
library(ROCR)
library(sciplot)
library(ggplot2)
library(pdp)
library(PresenceAbsence)
library(tidyr)
library(viridis)
library(caper)
###to install ggtree, need to first install BiocManager:
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("ggtree")
# library(phylofactor)
library(ggtree)
library(treeio)
library(caret) 
library(InformationValue)
library(mgcv) #for beta regression on performance metrics
# library(cowplot) #for combining plots of best.iter

# Clean environment
rm(list=ls()) 
graphics.off()

# Set working directory
setwd("/Users/katietseng/Library/CloudStorage/OneDrive-WashingtonStateUniversity(email.wsu.edu)/Fernandez Lab/Projects (Active)/OPV Host Prediction/GitHub/PoxHost")

```

### *Reclassify (drop) feline poxvirus as cowpox virus ###

```{r brt_}

# Load data and clean environment
load("data/linkpred_cleandata.RData")
data <- linkdata
rm(linkdata)

# Identify known interactions between feline poxvirus and mammal genera
data[data$virus=="feline poxvirus ita2_bc" & data$link==1, c('source','genus')]

# Identify whether known interaction between cowpox virus and Homo exists
data[data$virus=="cowpox virus" & data$link==1 & data$genus=="Homo", c('source','genus')]

# Drop all feline poxvirus pairings since cowpoxvirus-homo link is already in our dataset
nrow(data[data$virus=="feline poxvirus ita2_bc",])
data <- data[data$virus!="feline poxvirus ita2_bc",]

```

### *Create variables of taxonomic family as predictors for the model*

```{r brt_taxo}

# Calculate number of pseudoabsences
length(which(data$link==0))

# Ensure all accessory gene variables are numeric
PC_columns <- colnames(data[which(grepl("PC",names(data)))])
data[,c(PC_columns)] <- lapply(data[c(PC_columns)],as.numeric)
str(data)

# Make binary variables for each taxonomic family; remove any duplicates
dums=dummy_cols(data["fam"])
dums=dums[!duplicated(dums$fam),]

# Ensure all family vars are factor
for(i in 1:ncol(dums)){
  dums[,i]=factor(dums[,i])
}

# Merge family taxa variables with dataset as predictors
data=merge(data,dums,by="fam",all.x=T)

# Drop unnecessary columns and clean environment
data$traitname=NULL
rm(dums, i, PC_columns)

```

### *Assess variation and availability of data*
We explore the variation and availability of our data and drop variables if more than 40% of observations are missing/NA. We also remove variables that are not needed for the BRT analysis and save the dataset as *data_LinkBRT.RData*.
Figure: *histogram_trait_coverage.png*
Table: *table_trait_coverage.csv*

```{r brt_var}

# Create mode function where for each variable, we extract the frequency of the most frequently occurring value for that variable and divide it by the number of non-NA elements in that variable
mode.prop <- function(x) {                
  ux <- unique(x[is.na(x)==FALSE])        # creates array of unique values
  tab <- tabulate(match(na.omit(x), ux))  # creates array of the frequency (number of times) a unique value appears in a column 
  max(tab)/length(x[is.na(x)==FALSE])     # max-frequency / number of elements in each column that are not NA
}

# Assess variation across columns (2 indicates columns)
vars=data.frame(apply(data,2,function(x) mode.prop(x)),
                apply(data,2,function(x) length(unique(x))))    # number of unique elements in each column

# Get names
vars$variables=rownames(vars)
names(vars)=c("var","uniq","column")

# # Round values
# vars$var=round(vars$var,2)

# Label variables "cut" if homogeneous (100%)
vars$keep=ifelse(vars$var<1,"keep","cut")
vars$keep=ifelse(vars$column%in%c('fam','source','sequence','link','virus','genus','ord','gtip','treename','cites','ed_equal'),'keep',vars$keep) # ensures we keep these columns
vars=vars[order(vars$keep),]

# Trim (creates array of column names to cut and removes from df)
keeps=vars[-which(vars$keep=="cut"),]$column

# Drop if no variation
data=data[keeps]
rm(keeps,vars)

# Assess missing values
mval=data.frame(apply(data,2,function(x) length(x[!is.na(x)])/nrow(data))) # proportion of values that are not NA

# Get names
mval$variables=rownames(mval)
names(mval)=c("comp","column")

# Plot frequency distribution of coverage among traits
png("figures/other/linkpred/histogram_trait_coverage.png", width=4,height=4,units="in",res=600)
ggplot(mval[!mval$column%in%c("gen","treename","pcr","competence","tip.label","fam"),],
       aes(comp))+
  geom_histogram(bins=50)+
  geom_vline(xintercept=0.60,linetype=2,size=0.5)+
  theme_bw()+
  theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank())+
  theme(axis.title.x=element_text(margin=margin(t=10,r=0,b=0,l=0)))+
  theme(axis.title.y=element_text(margin=margin(t=0,r=10,b=0,l=0)))+
  labs(y="frequency",
       x="Trait coverage across mammal genera")+
  scale_x_continuous(labels = scales::percent)
dev.off()

# Label variables "cut" if >40% values are NA
mval$keep=ifelse(mval$comp>=0.60,"keep","cut")
table(mval$keep)
mval=mval[order(mval$keep),]

# Trim (creates array of column names to cut and removes from df)
keeps=mval[-which(mval$keep=="cut"),]$column

# Drop if not well represented
data=data[keeps]
rm(keeps,mval)

# Subset data to include only covariates
set <- subset(data,select=-c(fam, source, sequence, virus, genus, ord, gtip, treename))

#  Get trait coverage
trait_coverage=data.frame(apply(set,2,function(x) length(x[!is.na(x)])/nrow(set)))

# Rename and reorder columns
trait_coverage$variables=rownames(trait_coverage)
names(trait_coverage)=c("coverage","feature")
rownames(trait_coverage)=NULL
trait_coverage=trait_coverage[!trait_coverage$feature%in%c("pcr","competence"),]
trait_coverage <- subset(trait_coverage,select=c(feature,coverage))

# Save table
write.csv(trait_coverage, "figures/other/linkpred/table_trait_coverage.csv")

# Check that binary variables are numeric and not factor (except for fam vars)
str(set)

# Remove vars not needed for BRT analysis
data$source=NULL
data$sequence=NULL
# save(data, file='figures/other/linkpred/data_LinkBRT.RData')

# Clean environment
rm(keeps, mval, trait_coverage)
```

### *Alternative datasets Models 2 (excluding vaccinia virus sequences) and Model 3 (host traits only)*
For Model 2, we prepare an alternative dataset that excludes vaccinia virus sequences (host associations with vaccinia virus) from BRT analysis. For Model 3, we prepare an alternative dataset that excludes viral traits from BRT analysis and is trained only on host traits (includes vaccinia virus).

```{r brt_alt}

# Model 2: Subset data to exclude host-vaccinia virus links
# data <- data[data$virus!="vaccinia virus",]
# set <- subset(data,select=-c(fam, virus, genus, ord, gtip, treename))

# Model 3: Subset data to exclude viral traits
# We prepare an alternative dataset that excludes viral traits from BRT analysis so that model is trained only on host traits.
# 
# # Subset data to exclude viral traits
# data <- subset(data,select=-c(PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10))
# set <- subset(data,select=-c(fam, virus, genus, ord, gtip, treename))

```

### *Tuning to assess model performance for each combination of tuning parameters*
We create a hyperparameter 'grid' that represents different combinations of parameter values to which we tune the model. We then save a table of the resulting model performance measures for 'Model 1', where we predict on all possible host-virus combinations. If model tuning on the alternative dataset (see above), {r brt_alt}, we recommend you modify the results table name to *par...Model2.csv*.
Table: *par_tuning_data_summary_Model1.csv*

```{r brt_tuning}

# Hyperparameter tuning ifelse
#hok="ok"
hok="notok"
if(hok!="ok"){
  
  ## hyperparameter grid
  hgrid=expand.grid(n.trees=5000,                              #creates df from all combinations of factor vars (1*3*3*1*10=90 obs & 5 vars)
                    interaction.depth=c(2,3,4),
                    shrinkage=c(0.01,0.001,0.0005),
                    n.minobsinnode=4,
                    seed=seq(1,10,by=1))
  # hgrid=expand.grid(n.trees=500,                              #creates df from all combinations of factor vars (1*3*3*1*10=90 obs & 5 vars)
  #                   interaction.depth=c(2,3,4),
  #                   shrinkage=c(0.1,0.01,0.005),
  #                   n.minobsinnode=4,
  #                   seed=seq(1,10,by=1))
  # fix trees
  hgrid$n.trees=ifelse(hgrid$shrinkage<0.001,hgrid$n.trees*3,hgrid$n.trees)
  
  ## trees, depth, shrink, min, prop 
  hgrid$id=with(hgrid,paste(n.trees,interaction.depth,shrinkage,n.minobsinnode))   #creates var 'id' concatenating values from each of the specified columns in hgrid
  
  ## sort by id then seed
  hgrid=hgrid[order(hgrid$id,hgrid$seed),]
  
  ## now add rows
  hgrid$row=1:nrow(hgrid)                                        #adds var 'row' based on row number in hgrid
  
  ## factor id
  hgrid$id2=factor(as.numeric(factor(hgrid$id)))                 #creates 9-level factor var 'id2' 
  
  
  ## function to assess each hyperpar combination
  hfit=function(row,response){
    
    ## make new data
    ndata=set
    
    ## correct response
    ndata$response=ndata[response][,1]                           #creates var 'response'
    
    ## remove raw
    # ndata$pcr=NULL
    # ndata$competence=NULL
    ndata$link=NULL
    
    ## use rsample to split
    set.seed(hgrid$seed[row])                                    #sets seed value of 1-10
    split=initial_split(ndata,prop=0.7,strata="response")        #creates single binary split of data into training set and testing set, where 70% of data is retained for modeling/analysis and resampling is created within the 'response' var
    
    ## test and train
    dataTrain=training(split)
    dataTest=testing(split)
    
    ## yTest and yTrain
    yTrain=dataTrain$response                                    #create array of just response values from training and testing set
    yTest=dataTest$response
    
    ## BRT
    set.seed(1)
    gbmOut=gbm(response ~ . ,data=dataTrain,                     #y~x; gbmOut contains list of 29 elements including train.error and valid.error referenced later in gbm.perf()
               n.trees=hgrid$n.trees[row],                       #total number of trees to fit (number of iterations; default is 100)
               distribution="bernoulli",
               shrinkage=hgrid$shrinkage[row],                   #equiv to learning rate or step-size reduction (smaller learning rate requires more trees, default is 0.1)
               interaction.depth=hgrid$interaction.depth[row],   #max depth of each tree (highest level of variable interactions allowed; default is 1)
               n.minobsinnode=hgrid$n.minobsinnode[row],         #min. number of obs in terminal nodes of trees
               cv.folds=5,class.stratify.cv=TRUE,                #no. of cross-val folds to perform; for cv.folds>1, returns estimate of generalization error in 'cv.error'
               bag.fraction=0.5,train.fraction=1,                #fraction of training set obs randomly selected to propose next tree in expansion - this is why we set.seed()
               n.cores=5,                                        #no. of CPU cores to use
               verbose=F)
               # par.details=(gbmParallel(num_threads=5)),
    
    ## performance
    par(mfrow=c(1,1),mar=c(4,4,1,1))                             #sets graphical parameters such that subsequent figure are drawn in a nr-by-nc array by mfrows respectively and gives the number of lines of margin to be specified on the four sides of the plot c(bottom, L, top, R) -> see 'best.iter' plot below 
    best.iter=gbm.perf(gbmOut,method="cv")                       #estimates optimal number of boosting iterations and plots 'training.error' performance measure; cv method extracts this optimal number using cross-validation
    
    ## predict with test data
    preds=predict(gbmOut,dataTest,n.trees=best.iter,type="response")  #number of trees based on the optimal number of boosting iterations as set above (5,352)
    
    ## known
    result=dataTest$response
    
    # ##estimate threshold value for classification of predicted probability
    # #library(pROC)
    # analysis <- roc(result,preds)  #roc([actual values],[predicted values])
    # e <- cbind(analysis$thresholds,analysis$sensitivities+analysis$specificities) #pulls each array and binds them into dataframe: 1st column are thresholds, 2nd column are sensitivities + specificities
    # 
    # ##optimum threshold value
    # opt_t <- subset(e,e[,2]==max(e[,2]))[,1] #subsets dataframe and returns the max (sens+spec) value of 2nd column of e 
    # #threshold<-opt_t #set as threshold value
    # #threshold = 0.2
    
    ## sensitivity and specificity                              #e.g., test run produced sensitivity of 0 b/c no predictedScores were > 0.5; and specificity of 1 b/c all predictedScores were <0.5
    sen=InformationValue::sensitivity(result,preds)              #calculates sensitivity (# of obs with event AND predicted to have event, divided by # of obs w/ event) for a given logit model where input is the actual binary flag (as numerica vector) for the response variable and the predicted probability scores for each observation; if predicted value is above the threshold (defaults to 0.5), it will be considered an event (1) or else a non-event (0)
    spec=InformationValue::specificity(result,preds)             #calculates specificity (# of obs w/o event AND predicted to not have event, divided by # of obs w/o event)  
    
    ## AUC on train
    auc_train=gbm.roc.area(yTrain,predict(gbmOut,dataTrain,n.trees=best.iter,type="response"))   #compute Information Retrieval measures for pairwise loss for a single group, where input is the observed value and the predicted value
    
    ## AUC on test
    auc_test=gbm.roc.area(yTest,predict(gbmOut,dataTest,n.trees=best.iter,type="response"))
    
    ## print
    print(paste("hpar row ",row," done; test AUC is ",auc_test,sep=""))  #prints "hpar row [x] done; test AUC is []"
    
    ## save outputs
    return(list(best=best.iter,                    #saves optimal number of iterations, AUC on training set, AUC on testing set, specificity, sensitivity, and row number as a list
                trainAUC=auc_train,
                testAUC=auc_test,
                spec=spec,
                sen=sen,
                wrow=row))
  }
  
  ## run the function for link
  hpars=lapply(1:nrow(hgrid),function(x) hfit(x,response="link"))
  
  ## get results
  hresults=data.frame(sapply(hpars,function(x) x$trainAUC),
                      sapply(hpars,function(x) x$testAUC),
                      sapply(hpars,function(x) x$spec),
                      sapply(hpars,function(x) x$sen),
                      sapply(hpars,function(x) x$wrow),
                      sapply(hpars,function(x) x$best))
  names(hresults)=c("trainAUC","testAUC",
                    "spec","sen","row","best")
  
  ## combine and save
  hsearch=merge(hresults,hgrid,by="row")
  
  # ## save
  # hsearch$type="PCR"
  
  # ## rerun the function for competence
  # hpars=lapply(1:nrow(hgrid),function(x) hfit(x,response="competence"))
  # 
  # ## get results
  # hresults=data.frame(sapply(hpars,function(x) x$trainAUC),
  #                     sapply(hpars,function(x) x$testAUC),
  #                     sapply(hpars,function(x) x$spec),
  #                     sapply(hpars,function(x) x$sen),
  #                     sapply(hpars,function(x) x$wrow),
  #                     sapply(hpars,function(x) x$best))
  # names(hresults)=c("trainAUC","testAUC",
  #                   "spec","sen","row","best")
  # 
  # ## combine and save
  # csearch=merge(hresults,hgrid,by="row")
  # 
  # ## assign data type
  # csearch$type="competence"
  # 
  # ## combine
  # search=rbind.data.frame(csearch,hsearch)
  # search$type=factor(search$type,levels=c("PCR","competence"))

  search=hsearch
  
  ## export
  write.csv(search,"figures/other/linkpred/par_tuning_data_summary_Model1.csv")
  # write.csv(search,"Output/par_tuning_data_summary_Model2.csv")

}else{
  
  ## 
  search=read.csv("figures/other/linkpred/par_tuning_data_summary_Model1.csv")
  # search=read.csv("Output/par_tuning_data_summary_Model2.csv")

}

```

### *Assess model tuning results*
We fit a beta regression model using mgcv::gam to explore the main interaction effects of tuning parameters, interaction depth and shrinkage, on performance metrics (AUC, sensitivity, and specificity). This model is appropriate for analyzing continuous response variables bounded b/w 0-1 (i.e,, beta distribution). Using ANOVA, we determine whether the coefficients b/w the two interactions depths, the two shrinkage rates, and the four possible interactions b/w them are significantly different. We then plot the performance of BRTs based on various parameter combinations. Finally, we assess the distribution of best.iter (optimal # of iterations) to determine the max # of trees to use for model training.
Figure: *boxplot_brt_tuning.png*

```{r brt_tuning_results}

search=read.csv("figures/other/linkpred/par_tuning_data_summary_Model1.csv")
# search=read.csv("figures/other/linkpred/par_tuning_data_summary_Model1.csv")
# search=read.csv("figures/other/linkpred/par_tuning_data_summary_Model2.csv")

# Convert parameters to factor and relabel values
search$shrinkage=factor(search$shrinkage)
lvl=rev(sort(unique(search$shrinkage)))  #sorts unique shrinkage par from large to small
search$shrinkage=factor(search$shrinkage,levels=lvl); rm(lvl)  #applies as factor
search$interaction.depth=factor(search$interaction.depth)

# Fit beta regression model to the test AUCs; ANOVA
mod=gam(testAUC~interaction.depth*shrinkage,  
        data=search,method="REML",family=betar)
anova(mod)

# Fit beta regression model to the sensitivities; ANOVA
mod=gam(sen~interaction.depth*shrinkage,
        data=search,method="REML",family=betar)
anova(mod)

# Fit beta regression model to the specificities; ANOVA
mod=gam(spec~interaction.depth*shrinkage,
        data=search,method="REML",family=betar)
anova(mod)

# To plot model tuning performance, transform dataframe from wide to long
search2=gather(search,measure,value,testAUC:sen)

# Relabel values and convert to factor 
search2$measure=plyr::revalue(search2$measure,
                              c("sen"="sensitivity",  
                                "spec"="specificity",
                                "testAUC"="test AUC"))
search2$measure=factor(search2$measure,
                       levels=c("test AUC","sensitivity","specificity"))

# Boxplot performance of model tuning w/ various parameter combinations
png("figures/supplementary/s11fig_boxplot_brt_tuning_link.png",width=3.5,height=8,units="in",res=600)
set.seed(1)
ggplot(search2,aes(shrinkage,value,
                   colour=interaction.depth,fill=interaction.depth))+
  geom_boxplot(alpha=0.25)+
  geom_point(alpha=0.75,
             position = position_jitterdodge(dodge.width=0.75))+
  theme_bw()+
  theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank())+
  theme(axis.title.x=element_text(margin=margin(t=10,r=0,b=0,l=0)))+
  theme(axis.title.y=element_text(margin=margin(t=0,r=10,b=0,l=0)))+
  facet_grid(search2$measure,scales="free_y",switch="y")+
  theme(strip.placement="outside",
        strip.background=element_blank())+
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=12),
        strip.text=element_text(size=12))+
  theme(legend.position="top")+
  scale_color_viridis(discrete=TRUE,option="D")+
  scale_fill_viridis(discrete=TRUE,option="D")+
  guides(colour=guide_legend(title="interaction depth"),
         fill=guide_legend(title="interaction depth"))+
  labs(y=NULL,
       x="learning rate")+
  scale_y_continuous(n.breaks=4)
dev.off()

# To determine optimal parameters for model training, subset tuning results by number of trees
search_nt5000 <- search[search$n.trees==5000,]
search_nt15000 <- search[search$n.trees==15000,]
search_nt5000_sh0.01 <- search_nt5000[search_nt5000$shrinkage==0.010,]  #subset models with shrinkage==0.010

# Plot best.iter to see max number of trees to include
p1 <- search_nt5000 %>%
  # ggplot( aes(x=best, fill=type)) +
  ggplot( aes(x=best)) +
  geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
  scale_fill_manual(values=c("#69b3a2", "#404080")) +
  ggtitle("(A) ntrees=5000 and shrinkage=0.001")
p2 <- search_nt15000 %>%
  # ggplot( aes(x=best, fill=type)) +
  ggplot( aes(x=best)) +  
  geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
  scale_fill_manual(values=c("#69b3a2", "#404080"))  +
  ggtitle("(B) ntrees=15000 and shrinkage=0.001")
p3 <- search_nt5000_sh0.01 %>%
  # ggplot( aes(x=best, fill=type)) +
  ggplot( aes(x=best)) +
  geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
  scale_fill_manual(values=c("#69b3a2", "#404080")) +
  ggtitle("(C) ntrees=5000 and shrinkage=0.01")

# Combine plots and save to output - commenting out for HPC due to issues installing cowplot
# png("Output/histogram_brt_tuning_bestiter.png",width=12,height=4,units="in",res=600)
# plot_grid(p1,p2,p3) +
#   ggtitle("Histogram of best iteration")
# dev.off()

# Clean
rm(search,search2,hok,mod,search_nt5000,search_nt15000,search_nt5000_sh0.01,p1,p2,p3)

```

### *BRT function for applying across multiple data partitions*
We create our BRT function for model training, where we apply the function across multiple data partitions.

```{r brt_partition}

# BRT function to use different data partitions
brt_part=function(seed,response){
  
  ## Make new dataset
  ndata=set
  
  ## Correct response variable
  ndata$response=ndata[response][,1]
  
  ## Remove raw response variable
    ndata$link=NULL

  ## For BRT where cites is the response variable...
  if(response=="cites"){
    
    ## We add 1 to cites if cites equals 0
    ndata$cites=ifelse(ndata$cites==0,1,ndata$cites)
    
  }else{
    
    ndata=ndata
    
  }
  
  ## Use rsample package to split data
  set.seed(seed)
  split=initial_split(ndata,prop=0.7,strata="response")
  
  ## Create test and train datasets
  dataTrain=training(split)
  dataTest=testing(split)
  
  ## Get response variable for test dataset and train dataset
  yTrain=dataTrain$response
  yTest=dataTest$response
  
  ## Save distribution
  dist=ifelse(response=="cites","poisson","bernoulli")
  
  ## Save number of trees based on previous plots of optimal iterations
  nt=ifelse(response=="cites",10000,
     ifelse(response=="link",4500,5000)) #see plots of best.iter 
  
  ## Run BRTs using gbm package
  set.seed(1)
  gbmOut=gbm(response ~ . ,data=dataTrain,
             n.trees=nt,
             distribution=dist,
             shrinkage=0.01, #see plots of best.iter 
             interaction.depth=3,
             n.minobsinnode=4,
             cv.folds=5,class.stratify.cv=TRUE,
             bag.fraction=0.5,train.fraction=1,
             n.cores=5,
             verbose=F)
            # par.details=(gbmParallel(num_threads=5)),

  ## Get optimal number of iterations using gbm.perf & set plotting parameters for permance chart generated by gbm.perf
  par(mfrow=c(1,1),mar=c(4,4,1,1))                         
  best.iter=gbm.perf(gbmOut,method="cv")  #estimates optimal number of boosting iterations for a gbm object                 
  
  ## Predict with test data, applying the optimal number of iterations as n.trees
  preds=predict(gbmOut,dataTest,n.trees=best.iter,type="response")
  
  ## Save known associations
  result=dataTest$response
  
  ## Get sensitivity and specificity
  sen=InformationValue::sensitivity(result,preds)
  spec=InformationValue::specificity(result,preds)
  
  ## Get AUC from model training
  auc_train=gbm.roc.area(yTrain,predict(gbmOut,dataTrain,n.trees=best.iter,type="response"))
  
  ## Get AUC from model test
  auc_test=gbm.roc.area(yTest,predict(gbmOut,dataTest,n.trees=best.iter,type="response"))
  
  ## Skip if poisson
  if(response=="cites"){
    
    perf=NA
    
  }else{
    
    ## Inner loop if yTest is all 0
    if(var(yTest)==0){
      
      perf=NA
    }else{
      
      ## To construct an ROC curve, create a prediction object using the predicted probabilities and the true class labels (known responses)
      pr=prediction(preds,dataTest$response)   
      
      ## Next, calculate the desired performance measures specified by the 'measures' argument
      perf=performance(pr,measure="tpr",x.measure="fpr")         #pr=prediction object; measure=performance measure for evaluation; x.measure=second perf measure (2-D)
      
      ## Create a dataframe of those performance values
      perf=data.frame(perf@x.values,perf@y.values)
      
      
      ## Rename columns
      names(perf)=c("fpr","tpr")
      
      ## Add seed
      perf$seed=seed
      
    }
  }
  
  ## Get relative importance
  bars=summary(gbmOut,n.trees=best.iter,plotit=F)
  bars$rel.inf=round(bars$rel.inf,2)

  ## Predict with cites
  preds=predict(gbmOut,data,n.trees=best.iter,type="response")
  #pred_data=data[c("gtip",'treename',"fam","ord","pcr","competence")]
  pred_data=data[c("virus","gtip",'treename',"fam","ord","link")]
  pred_data$pred=preds
  pred_data$type=response
  
  ## Predict with mean cites
  pdata=data
  pdata$cites=mean(pdata$cites)
  pred_data$cpred=predict(gbmOut,pdata,n.trees=best.iter,type="response")
  
  ## Sort by decreasing predicted probability
  pred_data=pred_data[order(pred_data$pred,decreasing=T),]
  
  ## Print
  print(paste("BRT ",seed," done; test AUC = ",auc_test,sep=""))
  
  ## Save outputs
  return(list(mod=gbmOut,
              best=best.iter,
              trainAUC=auc_train,
              testAUC=auc_test,
              spec=spec,
              sen=sen,
              roc=perf,
              rinf=bars,
              predict=pred_data,
              traindata=dataTrain,
              testdata=dataTest,
              seed=seed))
}

```

### *Apply BRT function across 100 partitions to generate ensemble*

```{r brt_ensemble}

# Apply across 100 splits each
# smax=101
smax=100
brts=lapply(1:smax,function(x) brt_part(seed=x,response="link"))

# Run wos brts
pm_brts=lapply(1:(smax),function(x) brt_part(seed=x,response="cites"))

# Save results to data folder
save(brts,pm_brts,file="data/brts_Model1.RData")
# save(brts,pm_brts,file="data/brts_Model2.RData")
# save(brts,pm_brts,file="data/brts_Model3.RData")


# # If running on HPC, save results to Output folder
# save(brts,pm_brts,file="Output/brts_Model1.RData")
# save(brts,pm_brts,file="Output/brts_Model2.RData")
# save(brts,pm_brts,file="Output/brts_Model3.RData")

```